---
title: "Final-Project"
author: "Barbarit Clara and Duyck Louna"
format:
  html: default
  pdf: default
editor: visual
---

# **Effect of Demographic structure on Housing prices - A municipality-level analysis in France from 2014 to 2024**

### Research Question: **How does the demographic composition of a municipality affect local housing prices?**

Understanding the determinants of housing prices has long been a central topic in urban economics. While demand-side factors such as income, interest rates, or credit conditions are often highlighted, demographic composition represents a less explored dimension. Housing demand is affected by demographic structure through household formation, age distribution, family size, and socioeconomic characteristics, which shape the type, size, and location of residences required by local populations. The objective of this project is therefore to study how changes in the demographic structure of French municipalities between 2014 and 2024 have influenced local housing prices.

France is a relevant case study for answering our research question because it displays a wide diversity of spatial and demographic profiles. The French real estate market has undergone significant changes over the past decade, including rising average prices, varying transaction volumes, and growing spatial inequalities. Studying the link between demographic trends and housing prices over a ten-year period can therefore shed light on the mechanisms through which population structure influences local housing markets.

To explore this question, the project combines several datasets. Housing indicators are available at the municipal level and on an annual basis for the period 2014-2024. Demographic data come from French censuses (2011, 2016 and 2022), which provide detailed information on age structure, household size, employment categories, and other population characteristics. Although census variables are only available for a few reference years, they still enable us to capture the significant changes that have occurred in the demographic composition during the period. Working with panel data allows us to track each municipality over time and to observe how changes in its demographic profile are linked to variations in housing prices. The econometric analysis will rely on fixed-effects models, which help isolate the effect of demographic variables by controlling for characteristics specific to each municipality that do not change over time (geography, local amenities, or long-run development patterns). Although causal inference is challenging due to potential reverse causality (high housing prices may attract or deter certain demographic groups), the panel structure helps reduce this concern by observing demographic changes before and after price movements. Additional strategies may also be considered.

Our project is structured as follows. First, we present a descriptive analysis of the main variables in our datasets, using graphical and spatial visualisations to highlight key trends in both housing prices and demographic composition. Then we develop an econometric framework in order to isolate the effect of demographic structure on real estate prices. Finally, we address potential identification challenges, such as reverse causality and omitted variable bias, and discuss how our empirical strategy attempts to mitigate them.

# Libraries

```{r, warning=FALSE, message=FALSE}
here::i_am("BARBARIT-DUYCK-Project.Rproj")
library(here)
library(dplyr)
library(tidyr)
library(ggplot2)
library(vroom)
library(purrr)
library(stringr)
library(sf)
```

# Link to the github project

<https://github.com/lounaduyck-afk/BARBARIT-DUYCK-Project.git>

# Loading and describing the data

## 1) Residential property market data

<https://www.data.gouv.fr/datasets/indicateurs-immobiliers-par-commune-et-par-annee-prix-et-volumes-sur-la-periode-2014-2024>

From this website, we downloaded 11 csv files, each file corresponding to a specific year from 2014 to 2024. Each file gathers aggregate variables related to residential property market at the municipality scale.

The data have been collected by the government, from 2014 to 2024. The latest update was on the 7th of July 2025.

### Methodology

Data come from the treatment of another datagouv database regarding the Request for Property Values. From this initial database, they took only single-sale transfers, for which the values lied between 15,000€ and 10,000,000€, the surface between 10m² and 250m² for apartments, and between 10m² and 400m² for houses.

The population is the residential property transactions in Mainland France. The transactions have been aggregated at the municipalities level, which means that each line corresponds to a municipality for a unique year. Each municipality is identified by its INSEE code.

It is important to note that this database covers the whole of metropolitan France, with the exception of the departments of Bas-Rhin, Haut-Rhin, and Moselle, where the relevant data fall under the *livre foncier* (land registry) system.

One of the main strengths of this database is that it provides a common reference framework over a near-national geographical scope, making it possible to compare territories with one another, to understand their specific characteristics, and to produce studies and indicators based on homogeneous methods.

### Data merging

```{r, warning=FALSE, message=FALSE}
DVF2014 <- vroom(here("Data","dvf2014.csv")) |>
  rename(PropMaison = propmaison,
         PropAppart = propappart)

DVF2015 <- vroom(here("Data","dvf2015.csv")) |>
  rename(PropMaison = propmaison,
         PropAppart = propappart)

DVF2016 <- vroom(here("Data","dvf2016.csv")) |>
  rename(PropMaison = propmaison,
         PropAppart = propappart)

DVF2017 <- vroom(here("Data","dvf2017.csv")) |>
  rename(PropMaison = propmaison,
         PropAppart = propappart)

DVF2018 <- vroom(here("Data","dvf2018.csv")) %>%
  rename(PropMaison = propmaison,
         PropAppart = propappart)

DVF2019 <- vroom(here("Data","dvf2019.csv")) %>%
  rename(PropMaison = propmaison,
         PropAppart = propappart)

DVF2020 <- vroom(here("Data","dvf2020.csv")) |>
  rename(PropMaison = propmaison,
         PropAppart = propappart)

DVF2021 <- vroom(here("Data","dvf2021.csv")) |>
  rename(PropMaison = propmaison,
         PropAppart = propappart)

DVF2022 <- vroom(here("Data","dvf2022.csv")) |>
  rename(PropMaison = propmaison,
         PropAppart = propappart)

DVF2023 <- vroom(here("Data","dvf2023.csv")) |>
  rename(PropMaison = propmaison,
         PropAppart = propappart)

DVF2024 <- vroom(here("Data","dvf2024.csv")) |>
  rename(Annee = annee,
         Nb_mutations = nb_mutations)
```

```{r, warning=FALSE, message=FALSE}
DVF_panel <- bind_rows(
  DVF2014, DVF2015, DVF2016, DVF2017, DVF2018,
  DVF2019, DVF2020, DVF2021, DVF2022, DVF2023, DVF2024) |>
  select(-1)
```

Explanations: after standardizing the labels, we merge our data using bind_rows in order to get panel data.

### Changing the columns type

```{r}
DVF_panel <- DVF_panel |>
  mutate(Annee = as.integer(Annee))
```

### Data descriptions

#### Basic Information

```{r, warning=FALSE, message=FALSE}
nb_columns <- ncol(DVF_panel)

nb_rows <- nrow(DVF_panel)

obs_by_year <- DVF_panel |>
  group_by(Annee) |>
  summarise(n_obs = n()) |>
  arrange(Annee)

summary_table <- tibble(
  Type = c("Columns", "Total Rows", paste("Observations", obs_by_year$Annee, sep=" ")),
  Number = c(nb_columns, nb_rows, obs_by_year$n_obs)
)

summary_table |>
  knitr::kable()
```

#### Variables descriptions

-   INSEE code: a qualitative variable which is unique for each municipality
-   Year
-   The number of transfers: quantitative variable which is the sum of the house sales and flat sales
-   The number of house sales: quantitative variable
-   The number of flat sales: quantitative variable
-   The proportion of house sales in percentage: quantitative variable
-   The proportion of flat sales in percentage: quantitative variable
-   The average price of properties sold in €: quantitative variable
-   The average price per square metre of properties sold in €: quantitative variable
-   The average surface area of properties sold in square metre: quantitative variable

## 2) Population evolution and stucture by municipalities

<https://www.insee.fr/fr/statistiques/8581696?sommaire=8581933>

From this website, we downloaded csv file corresponding to "Évolution et structure de la population en 2022 - Commune - France hors Mayotte"

We have population statistics in 2011, 2016 and 2022 for each municipality in France, except Mayotte, corresponding to the January 2025 geography.

They have been collected by the INSEE during the census.

The data cover population distribution by sex, age, occupations but also housing conditions and relocation.

```{r, warning=FALSE, message=FALSE}
Population <- vroom(here("Data","base-cc-evol-struct-pop-2022.csv"))
Labels <- vroom(here("Data","meta_base-cc-evol-struct-pop-2022.csv"))
```

### Data cleaning

#### Standardizing the labels

```{r, warning=FALSE, message=FALSE}
new_names <- setNames(Labels$COD_VAR, Labels$LIB_VAR)
Population <- Population |>
  rename(!!!new_names)
Population <- Population |>
  rename(INSEE_COM = `Arrondissement municipal`, "Pop 15 ans ou plus Autres personnes sans activité professionnelle en 2016 (compl)" = `Pop 15 ans ou plus Autres en 2016 (compl)`, "Pop 15 ans ou plus Hommes Autres personnes sans activité professionnelle en 2016 (compl)" = `Pop 15 ans ou plus Hommes Autres en 2016 (compl)`, "Pop 15 ans ou plus Femmes Autres personnes sans activité professionnelle en 2016 (compl)
" = `Pop 15 ans ou plus Femmes Autres en 2016 (compl)`, "Pop 15-24 ans Autres personnes sans activité professionnelle en 2016 (compl)
" = `Pop 15-24 ans Autres en 2016 (compl)`, "Pop 25-54 ans Autres personnes sans activité professionnelle en 2016 (compl)
" = `Pop 25-54 ans Autres en 2016 (compl)`, "Pop 55 ans ou plus Autres personnes sans activité professionnelle en 2016 (compl)" = `Pop 55 ans ou plus Autres en 2016 (compl)`)
```

Explanations: When we downloaded the data sets, we had two csv files, one with the data we want to look at, in which the labels had a specific code, and the other with the labels' code and their meaning. Then, we changed the code labels of our data set of interest to get their meaning.

### Modificating the structure of the dataset

```{r, warning=FALSE, message=FALSE}
Population <- Population |>
  pivot_longer(
    cols = -INSEE_COM,
    names_to = "temp",
    values_to = "value"
  ) |>
  mutate(
    Annee = str_extract(temp, "\\d{4}"),                
    Variable = str_remove(temp, " en \\d{4}.*$")               
  ) |>
  select(INSEE_COM, Annee, Variable, value) |>
  pivot_wider(
    names_from = Variable,
    values_from = value
  ) |>
  arrange(INSEE_COM, Annee)
```

Explanations:

Initially, we had one row for each municipality and the variables for each year. For example, we have 3 columns related to the number of men in the municipality, one for 2022, 2016 and 2011 ("Pop Homme en 2022", "Pop Homme en 2016", "Pop Homme en 2011"). Therefore, in order to get a panel data structure as the DVF data set we uploaded in the first part, we used at first pivot longer and we extract the year in each initial label. Then we create an additional column "Annee" and we used pivot wider to have the panel structure.

### Changing the columns type

```{r}
Population <- Population |>
  mutate(Annee = as.integer(Annee))
```

### Data descriptions

#### Basic Information

```{r, warning=FALSE, message=FALSE}
nb_columns <- ncol(Population)

nb_rows <- nrow(Population)

obs_by_year <- Population |>
  group_by(Annee) |>
  summarise(n_obs = n()) |>
  arrange(Annee)

summary_table <- tibble(
  Type = c("Columns", "Total Rows", paste("Observations", obs_by_year$Annee, sep=" ")),
  Number = c(nb_columns, nb_rows, obs_by_year$n_obs)
)

summary_table |>
  knitr::kable()
```

#### Variables description

-   Population: quantitative variable - number of residents in each municipality
-   Per age range population (0-14, 15-29, 30-44, 45-59, 60-74, 75-89, 90+)
-   Men population
-   Per age range men population (0-14, 15-29, 30-44, 45-59, 60-74, 75-89, 90+)
-   Women population
-   Per age range women population (0-14, 15-29, 30-44, 45-59, 60-74, 75-89, 90+)
-   Number of residents who live in the municipality for more than 1 year and who live in the same accommodation than 1 year ago
-   Number of residents who live in the municipality for more than 1 year and who live in a different accommodation than 1 year ago
-   Number of residents who lived outside the municipality 1 year ago (in a different department, in a different region, in Mainland France or in DOM) and their distribution by age and by sex.
-   Number of residents in each CSP for every age range and sex.

## 3) Unemployment

<https://www.observatoire-des-territoires.gouv.fr/taux-de-chomage-des-15-64-ans-rp>

From this website, we downloaded the file corresponding to "Taux de chômage des 15-64 ans", which was initially in the xlsx format. We therefore opened it on Excel and stored it as a csv file.

The data have been collected by the Insee, using the census in 2011, 2012, 2016 and 2022.

The data set gather, for each municipality and for each year of census, the total number of unemployed residents, the female unemployed residents and the male unemployed residents.

```{r, warning=FALSE, message=FALSE}
Unemployment <- vroom(here("Data","indic_sex_rp.csv"),
  delim = ";",
  skip = 4,
  col_names = TRUE,
  col_types = cols(
    codgeo = col_character(),
    libgeo = col_character(),
    an = col_integer(),
    sexe = col_factor(),
    nb_chom1564 = col_integer())) |>
  rename(INSEE_COM = codgeo,
         Annee = an,
         Chomeur = nb_chom1564)
```

```{r}
Unemployment |>
  filter(Annee == 2012) |>
  summarise(na_chomeur = sum(is.na(Chomeur)),
    obs = n()) |>
  knitr::kable()
```

Since all the observations in 2012 are NA, we apply a filter to remove them.

```{r}
Unemployment <- Unemployment |>
  filter(!Annee==2012)
```

### Modificating the structure of the dataset

```{r}
Unemployment <- Unemployment |>
  mutate(sexe = case_when(
    sexe == "F" ~ "chomeur_femme",
    sexe == "H" ~ "chomeur_homme",
    sexe == "T" ~ "chomeur_total"
  )) |>
  pivot_wider(
    names_from = sexe,
    values_from = Chomeur
  )
```

Initially, for each year, each municipality had three rows:

\- Total number of unemployed residents

\- The number of female unemployed residents

\- The number of male unemployed residents To get a consistent panel structure, we created additional columns, that specify to what each number refers, and we used pivot wider.

### Data description

#### Basic information

```{r, warning=FALSE, message=FALSE}
nb_columns <- ncol(Unemployment)

nb_rows <- nrow(Unemployment)

obs_by_year <- Unemployment |>
  group_by(Annee) |>
  summarise(n_obs = n()) |>
  arrange(Annee)

summary_table <- tibble(
  Type = c("Columns", "Total Rows", paste("Observations", obs_by_year$Annee, sep=" ")),
  Number = c(nb_columns, nb_rows, obs_by_year$n_obs)
)

summary_table |>
  knitr::kable()
```

#### Variables description

-   INSEE_COM: as the previous data bases, this one uses the Insee code to classify the municipality
-   libgeo: it corresponds to the name of the municipality associated with the INSEE code
-   Annee
-   Sex: coded as a factor, it expresses the gender category of the observation, distinguishing between females, males, and the total population
-   Chomeur: number of unemployed people (based on the INSEE's definition of unemployment)

# Merging the databases

## Code

```{r, warning=FALSE, message=FALSE}
DVFcensus <- DVF_panel |>
  filter(Annee %in% c(2022, 2016))

Population_filtered <- Population |>
  filter(Annee %in% c(2022, 2016))

Unemployment_filtered <- Unemployment |>
  filter(Annee%in%c(2016,2022))
```

```{r, warning=FALSE, message=FALSE}
Merge <- left_join(
  DVFcensus,
  Population_filtered,
  by = join_by(INSEE_COM, Annee)
)

Merge <- Merge |>
  left_join(
    Unemployment_filtered,
    by = join_by(INSEE_COM, Annee))
```

## Description

From the 3 data bases, we only kept data from 2016 and 2022, years of census and common to all bases. Using left join, because we wanted to keep all the observations we have in the initial panel set and match them with additional data we have from the two other data sets, we merged the data sets. Before merging, we ensured that the structure of the three data sets where coherent to be merged. We did this work in previous sections.

## Geographical data

​​<https://geoservices.ign.fr/adminexpress>

The ADMIN EXPRESS product line centralizes five related datasets: ADMIN EXPRESS, ADMIN EXPRESS COG, ADMIN EXPRESS COG CARTO, ADMIN EXPRESS COG CARTO Petite Échelle, and ADMIN EXPRESS COG CARTOPLUS PE. Together, these datasets provide detailed administrative boundary layers for France across multiple levels. They include the following geographic units:

-   Arrondissements (departmental and municipal)

-   Cantons

-   Administrative seats (chef-lieux) at various levels: arrondissement, municipal arrondissement, EPCI, canton, territorial collectivity, commune, associated or delegated commune, département, and region

-   Communes and associated or delegated communes

-   Départements

-   EPCIs (public inter-municipal cooperation bodies)

-   Regions

The main product, ADMIN EXPRESS, provides large-scale administrative boundaries and is updated monthly. An additional edition, ADMIN EXPRESS COG, is aligned with the Code Officiel Géographique (COG) published annually by INSEE.

This product line facilitates combining administrative boundaries with external datasets to build thematic representations of the French territory at different administrative levels. For our project, we mainly rely on the commune layer, as it gives the spatial unit necessary to merge geographic boundaries with demographic, employment, and real-estate datasets at the municipal level.

Here, we import both layers using `st_read()`, then standardize their INSEE municipality codes by trimming extra spaces and ensuring they always have five digits. This harmonization is essential for correctly joining the geographic layers with our DVF database, which also uses INSEE codes as identifiers.

```{r, warning=FALSE, message=FALSE}
gpkg_path <- "Data/ADE_4-0_GPKG_LAMB93_FXX-ED2025-10-15.gpkg"
communes_sf <- st_read(gpkg_path, layer = "chef_lieu_de_commune") |>
mutate(code_insee_de_la_commune = sprintf("%05s", trimws(code_insee_de_la_commune)))

communes_poly <- st_read(gpkg_path, layer = "commune") |>
mutate(code_insee = sprintf("%05s", trimws(code_insee)))
```

# Graphical representations

## 2024 (points)

```{r, warning=FALSE, message=FALSE}
df_2024 <- DVF_panel |>
filter(Annee == 2024)

carte_2024 <- communes_sf |>
left_join(df_2024, by = c("code_insee_de_la_commune" = "INSEE_COM"))

ggplot(carte_2024) +
geom_sf(aes(color = PrixMoyen), size = 1, alpha = 0.7) +
scale_color_viridis_c(trans = "log",direction = -1) +
labs(title = "Average Price per Municipality (2024)", color = "Average prices (€)") +
theme_minimal()
```

```{r, warning=FALSE, message=FALSE}
df_2014 <- DVF_panel |>
filter(Annee == 2014)

carte_2014 <- communes_sf |>
left_join(df_2014, by = c("code_insee_de_la_commune" = "INSEE_COM"))

ggplot(carte_2014) +
geom_sf(aes(color = PrixMoyen), size = 1, alpha = 0.7) +
scale_color_viridis_c(trans = "log", direction = -1) +
labs(title = "Average Price per Municipality (2014)", color = "Average Price (€)") +
theme_minimal()
```

```{r, warning=FALSE, message=FALSE}
departements_idf <- c("75","92","93","94","91","77","78","95")

communes_idf <- communes_poly |>
filter(substr(code_insee, 1, 2) %in% departements_idf)

carte_idf_2024 <- communes_idf |>
left_join(df_2024, by = c("code_insee" = "INSEE_COM"))

ggplot(carte_idf_2024) +
geom_sf(aes(fill = PrixMoyen), color = "white", size = 0.1) +
scale_fill_viridis_c(trans = "log", direction = -1) +
theme_minimal() +
labs(title = "Average Price — Île-de-France (2024)", fill = "Average Price (€)")
```

### Graphic of the evolution of the average property price

```{r, warning=FALSE, message=FALSE}
DVF_panel |>
  group_by(Annee) |>
  summarise(prix_moyen = mean(PrixMoyen, na.rm = TRUE)) |>
  ggplot(aes(x = Annee, y = prix_moyen)) +
  geom_line(linewidth = 1.2, color = "steelblue") +
  geom_point(size = 2, color = "steelblue") +
  theme_minimal() +
    labs(
    title = "Evolution of the Average Property Price",
    x = "Year",
    y = "Average Price (€)"
  )

```

### Graphic of the evolution of the average price of real estate per square meter

```{r, warning=FALSE, message=FALSE}
DVF_panel |>
  group_by(Annee) |>
  summarise(prix_moyen_m2 = mean(Prixm2Moyen, na.rm = TRUE)) |>
  ggplot(aes(x = Annee, y = prix_moyen_m2)) +
  geom_line(linewidth = 1.2, color = "steelblue") +
  geom_point(size = 2, color = "steelblue") +
  theme_minimal() +
    labs(
    title = "Evolution of the Average Property Price per square meter",
    x = "Year",
    y = "Average price per m² (€)"
  )
```

### Graphic of the evolution of the number of transactions

```{r, warning=FALSE, message=FALSE}
DVF_panel |>
  group_by(Annee) |>
  summarise(Total_mutations = sum(Nb_mutations, na.rm = TRUE)) |>
  ggplot(aes(x = Annee, y = Total_mutations)) +
  geom_line(color = "steelblue", size = 1.2) +
  geom_point(color = "red", size = 2) +
  labs(title = "Evolution of the number of real estate transactions in France (2014-2024)",
       x = "Year",
       y = "Number of transactions") +
  theme_minimal()
```

### Graphical representation of the Average property size over time

```{r, warning=FALSE, message=FALSE}
DVF_panel |>
  group_by(Annee) |>
  summarise(SurfaceMoy = mean(SurfaceMoy, na.rm = TRUE)) |>
  ggplot(aes(x = Annee, y = SurfaceMoy)) +
  geom_line(color = "darkred", size = 1.2) +
  geom_point(color = "red", size = 2) +
  labs(title = "Evolution of the average surface area of properties in France",
       x = "Year",
       y = "Average surface area (m²)") +
  theme_minimal()
```

### Distribution of houses vs. apartments by year

```{r, warning=FALSE, message=FALSE}
DVF_panel |>
  group_by(Annee) |>
  summarise(TotalMaisons = sum(NbMaisons, na.rm = TRUE),
            TotalApparts = sum(NbApparts, na.rm = TRUE)) |>
  pivot_longer(cols = c(TotalMaisons, TotalApparts),
               names_to = "TypeBien",
               values_to = "Nombre") %>%
  ggplot(aes(x = Annee, y = Nombre, fill = TypeBien)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Distribution of houses and apartments per year",
       x = "Year",
       y = "Number of transactions") +
  scale_fill_manual(values = c("TotalMaisons" = "orange", "TotalApparts" = "purple"),
                    labels = c("Maisons", "Appartements")) +
  theme_minimal()
```

```{r, warning=FALSE, message=FALSE}
DVF_panel |>
  group_by(Annee) |>
  summarise(
    TotalMaisons = sum(NbMaisons, na.rm = TRUE),
    TotalApparts = sum(NbApparts, na.rm = TRUE)
  ) |>
  mutate(
    Total = TotalMaisons + TotalApparts,
    ShareMaisons = TotalMaisons / Total,
    ShareApparts = TotalApparts / Total
  ) |>
  pivot_longer(
    cols = c(ShareMaisons, ShareApparts),
    names_to = "TypeBien",
    values_to = "Share"
  ) |>
  ggplot(aes(x = Annee, y = Share, fill = TypeBien)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Distribution of houses and apartments per year",
    x = "Year",
    y = "Share of transactions (%)"
  ) +
  scale_fill_manual(
    values = c("ShareMaisons" = "orange", "ShareApparts" = "purple"),
    labels = c("Houses", "Apartments")
  ) +
  theme_minimal()

```

## Comparison across the 10 largest French metropolitan areas

```{r, warning=FALSE, message=FALSE}
metro_labels <- tibble::tibble(
  INSEE_COM = c("75056", "69123", "33063", "13055", "31555",
                "06088", "44109", "34172", "67482", "59350"),
  Metro = c("Paris",
            "Lyon",
            "Bordeaux",
            "Marseille",
            "Toulouse",
            "Nice",
            "Nantes",
            "Montpellier",
            "Strasbourg",
            "Lille")
)
```

```{r, warning=FALSE, message=FALSE}
df_comp <- DVF_panel %>%
  filter(INSEE_COM %in% communes_compare) %>%
  left_join(metro_labels, by = "INSEE_COM")
```

```{r, warning=FALSE, message=FALSE}
ggplot(df_comp, aes(x = Annee, y = Prixm2Moyen, color = Metro)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2) +
  theme_minimal() +
  labs(
    title = "Comparison across the 10 largest French metropolitan areas (m2)",
    x = "Year",
    y = "Average property price (€)",
    color = "Metropolitan area"
  ) +
  scale_color_brewer(palette = "Set2")
```

```{r, warning=FALSE, message=FALSE}
df_comp %>%
  filter(!is.na(Nb_mutations)) %>%
  ggplot(aes(x = Annee, y = Nb_mutations, color = Metro)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.8) +
  theme_minimal() +
  labs(
    title = "Comparison of transaction volumes across the 10 French metropolitan areas",
    x = "Year",
    y = "Number of transactions",
    color = "Metropolitan area"
  ) +
scale_color_brewer(palette = "Set2")
```

## Demographic indicators

To go further in the data analysis process, we calculate several indicators.

-   **The upper socio-economic group rate**: percentage of managers and higher intellectual professions among the population of 15+ years old.

-   **The unemployment rate**: the percentage of unemployed people among the active population (between 20 years old and 64 years old).

    It should be noted that this ratio is approximate, as it does not take into account individuals aged 15 to 19. As we did not have the number of inhabitants for each age group in each municipality, we chose the age group closest to the working population.

-   **The relocation rate**: the percentage of people that initiate a transaction during the year while they were not resident in this municipality previously.

```{r} Merge <- Merge |>   mutate(     TauxCSPsup =`Pop 15 ans ou plus Cadres, Prof. intel. sup.`/ `Pop 15 ans ou plus`*100,     TauxChom = chomeur_total /(`Pop Hommes 20-64 ans`+`Pop Femmes 20-64 ans`)*100,     TauxDem = (1 - (`Pop 1 an ou plus habitant 1 an avt même logt`+`Pop 1 an ou plus habitant 1 an avt autre logt même commune`)/Population)*100)}
```

### The upper socio-economic group rate representation

```{r}
#Merging the geolocalisation dataset with the demographic indicators for each municipality
map_demo <- communes_sf |>
  left_join(
    Merge |> filter(Annee %in% c(2016, 2022)),
    by = c("code_insee_de_la_commune" = "INSEE_COM")) |>
  select(code_insee_de_la_commune, Annee, TauxCSPsup, TauxChom, TauxDem)
```

```{r}
map_demo |>
  filter(!is.na(TauxCSPsup)) |>
  ggplot() +
  geom_sf(aes(color = TauxCSPsup), size = 1, alpha = 0.7) +
  facet_wrap(~ Annee, ncol = 2, scales = "fixed") +
  scale_color_viridis_c(direction = -1) +
  labs(
    title = "CSP+ rate by municipalities",
    color = "Rate (%)") +
  theme_minimal()
```

### The unemployment rate representation

```{r}
map_demo |>
  filter(!is.na(TauxChom)) |>
  ggplot() +
  geom_sf(aes(color = TauxChom), size = 1, alpha = 0.7) +
  facet_wrap(~ Annee, ncol = 2, scales = "fixed") +
  scale_color_viridis_c(direction = -1) +
  labs(
    title = "Unemployment rate by municipalities",
    color = "Rate (%)"
  ) +
  theme_minimal()
```

### The relocation rate representation

```{r}
map_demo |>
  filter(!is.na(TauxDem)) |>
  ggplot() +
  geom_sf(aes(color = TauxDem), size = 1, alpha = 0.7) +
  facet_wrap(~ Annee, ncol = 2, scales = "fixed") +
  scale_color_viridis_c(direction = -1) +
  labs(
    title = "Relocation rate by municipalities",
    color = "Rate (%)"
  ) +
  theme_minimal()
```

### Analysis per departments

#### Distribution of socio-professional groups across departments

```{r}
csp_cols <- c(
  "Pop 15 ans ou plus Agriculteurs exploitants",
  "Pop 15 ans ou plus Artisans, Comm., Chefs entr.",
  "Pop 15 ans ou plus Cadres, Prof. intel. sup.",
  "Pop 15 ans ou plus Prof. intermédiaires ",
  "Pop 15 ans ou plus Employés",
  "Pop 15 ans ou plus Ouvriers",
  "Pop 15 ans ou plus Retraités",
  "Pop 15 ans ou plus Autres personnes sans activité professionnelle")

Merge |>
  filter(Annee %in% c(2016, 2022)) |>
  
  filter(!is.na(code_departement)) |>
  mutate(
    dep_label = paste0(code_departement, " – ", nom_departement),
    Annee = factor(Annee)
  ) |>
  pivot_longer(
    cols = all_of(csp_cols),
    names_to = "variable",
    values_to = "pop"
  ) |>
  mutate(
    CSP = str_remove(variable, "^Pop 15 ans ou plus\\s*"),
    CSP = str_squish(CSP)
  ) |>
  group_by(dep_label, Annee, CSP) |>
  summarise(pop = sum(pop, na.rm = TRUE), .groups = "drop") |>
  group_by(dep_label, Annee) |>
  mutate(part = pop / sum(pop)) |>
  ggplot(aes(x = Annee, y = part, fill = CSP)) +
  geom_col(width = 0.8) +
  coord_flip() +
  facet_wrap(~ dep_label, ncol = 6) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Distribution of socio-professional groups across departments.",
    subtitle = "Comparison between 2016 and 2022 (population aged 15 and over)",
    x = NULL,
    y = "Share",
    fill = "Socio-professional groups"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold", size = 9),
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(size = 8),
    panel.grid.major.y = element_blank()
  )
```

```{r}
# =========================
# 2022
# =========================
Merge |>
  filter(Annee == 2022, !is.na(code_departement)) |>
  group_by(code_departement, nom_departement) |>
  summarise(
    cadres = sum(`Pop 15 ans ou plus Cadres, Prof. intel. sup.`, na.rm = TRUE),
    denom  = sum(`Pop 15 ans ou plus`, na.rm = TRUE),
    TauxCSPsup = 100 * cadres / denom,
    .groups = "drop"
  ) |>
  filter(is.finite(TauxCSPsup), denom > 0) |>
  arrange(TauxCSPsup) |>
  (\(d) bind_cols(
    d |>
      slice_head(n = 10) |>
      transmute(`Lowest shares (2022)` = nom_departement,
                `CSP+ share` = round(TauxCSPsup, 2)),
    d |>
      slice_tail(n = 10) |>
      arrange(desc(TauxCSPsup)) |>
      transmute(`Highest shares (2022)` = nom_departement,
                `CSP+ share` = round(TauxCSPsup, 2))
  )) |>
  kable(
    digits = 2,
    caption = "CSP+ — Departments with the lowest and highest shares of executives (2022)"
  )

# =========================
# 2016
# =========================
Merge |>
  filter(Annee == 2016, !is.na(code_departement)) |>
  group_by(code_departement, nom_departement) |>
  summarise(
    cadres = sum(`Pop 15 ans ou plus Cadres, Prof. intel. sup.`, na.rm = TRUE),
    denom  = sum(`Pop 15 ans ou plus`, na.rm = TRUE),
    TauxCSPsup = 100 * cadres / denom,
    .groups = "drop"
  ) |>
  filter(is.finite(TauxCSPsup), denom > 0) |>
  arrange(TauxCSPsup) |>
  (\(d) bind_cols(
    d |>
      slice_head(n = 10) |>
      transmute(`Lowest shares (2016)` = nom_departement,
                `CSP+ share` = round(TauxCSPsup, 2)),
    d |>
      slice_tail(n = 10) |>
      arrange(desc(TauxCSPsup)) |>
      transmute(`Highest shares (2016)` = nom_departement,
                `CSP+ share` = round(TauxCSPsup, 2))
  )) |>
  kable(
    digits = 2,
    caption = "CSP+ — Departments with the lowest and highest shares of executives (2016)"
  )

```

#### Unemployment rate across departments

```{r}
chom_dep <- Merge |>
  filter(Annee %in% c(2016, 2022)) |>
  filter(!is.na(code_departement)) |>
  mutate(
    dep_label = paste0(code_departement, " – ", nom_departement),
    Annee = factor(Annee),
    pop2064 = `Pop Hommes 20-64 ans` + `Pop Femmes 20-64 ans`
  ) |>
  group_by(dep_label, Annee) |>
  summarise(
    chomeurs = sum(chomeur_total, na.rm = TRUE),
    pop2064  = sum(pop2064, na.rm = TRUE),
    taux_chom = 100 * chomeurs / pop2064,
    .groups = "drop"
  )


ggplot(chom_dep, aes(x = Annee, y = taux_chom, fill = Annee)) +
  geom_col(width = 0.8) +
  coord_flip() +
  facet_wrap(~ dep_label, ncol = 6) +
  scale_fill_manual(
    values = c(
      "2016" = "#66C2A5",
      "2022" = "#FC8D62"
    )
  ) +
  labs(
    title = "Unemployment Rate by Department",
    subtitle = "Comparison between 2016 and 2022 (ratio of unemployed persons to the population aged 20–64).",
    x = NULL,
    y = "Unemployment rate (%)",
    fill = "Year"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold", size = 9),
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(size = 8),
    panel.grid.major.y = element_blank()
  )
```

#### Relocation Rate across departments

```{r}
dem_dep <- Merge |>
  filter(Annee %in% c(2016, 2022)) |>
  filter(!is.na(code_departement)) |>
  mutate(
    dep_label = paste0(code_departement, " – ", nom_departement),
    Annee = factor(Annee),
    non_demenage = 
      `Pop 1 an ou plus habitant 1 an avt même logt` +
      `Pop 1 an ou plus habitant 1 an avt autre logt même commune`
  ) |>
  group_by(dep_label, Annee) |>
  summarise(
    non_demenage = sum(non_demenage, na.rm = TRUE),
    population   = sum(Population, na.rm = TRUE),
    taux_dem = 100 * (1 - non_demenage / population),
    .groups = "drop"
  )


ggplot(dem_dep, aes(x = Annee, y = taux_dem, fill = Annee)) +
  geom_col(width = 0.8) +
  coord_flip() +
  facet_wrap(~ dep_label, ncol = 6) +
  scale_fill_manual(
    values = c(
      "2016" = "#8DA0CB",  
      "2022" = "#E78AC3" 
    )
  ) +
  labs(
    title = "Residential Mobility Rate by Department",
    subtitle = "2016 vs 2022 Comparison (Resident Population)",
    x = NULL,
    y = "Relocation rate (%)",
    fill = "Year"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold", size = 9),
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(size = 8),
    panel.grid.major.y = element_blank()
  )
```
