---
title: "Final-Project"
author: "Barbarit Clara and Duyck Louna"
format:
  html: default
  pdf: default
editor: visual
---

# **Effect of Demographic structure on Housing prices - A municipality-level analysis in France from 2014 to 2024**

### Research Question: **How does the demographic composition of a municipality affect local housing prices?**

Understanding the determinants of housing prices has long been a central topic in urban economics. While demand-side factors such as income, interest rates, or credit conditions are often highlighted, demographic composition represents a less explored dimension. Housing demand is affected by demographic structure through household formation, age distribution, family size, and socioeconomic characteristics, which shape the type, size, and location of residences required by local populations. The objective of this project is therefore to study how changes in the demographic structure of French municipalities between 2014 and 2024 have influenced local housing prices.

France is a relevant case study for answering our research question because it displays a wide diversity of spatial and demographic profiles. The French real estate market has undergone significant changes over the past decade, including rising average prices, varying transaction volumes, and growing spatial inequalities. Studying the link between demographic trends and housing prices over a ten-year period can therefore shed light on the mechanisms through which population structure influences local housing markets.

To explore this question, the project combines several datasets. Housing indicators are available at the municipal level and on an annual basis for the period 2014-2024. Demographic data come from French censuses (2011, 2016 and 2022), which provide detailed information on age structure, household size, employment categories, and other population characteristics. Although census variables are only available for a few reference years, they still enable us to capture the significant changes that have occurred in the demographic composition during the period. Working with panel data allows us to track each municipality over time and to observe how changes in its demographic profile are linked to variations in housing prices. The econometric analysis will rely on fixed-effects models, which help isolate the effect of demographic variables by controlling for characteristics specific to each municipality that do not change over time (geography, local amenities, or long-run development patterns). Although causal inference is challenging due to potential reverse causality (high housing prices may attract or deter certain demographic groups), the panel structure helps reduce this concern by observing demographic changes before and after price movements. Additional strategies may also be considered.

Our project is structured as follows. First, we present a descriptive analysis of the main variables in our datasets, using graphical and spatial visualisations to highlight key trends in both housing prices and demographic composition. Then we develop an econometric framework in order to isolate the effect of demographic structure on real estate prices. Finally, we address potential identification challenges, such as reverse causality and omitted variable bias, and discuss how our empirical strategy attempts to mitigate them.

# Libraries

```{r, warning=FALSE, message=FALSE, echo=FALSE}
here::i_am("BARBARIT-DUYCK-Project.Rproj")
library(here)
library(dplyr)
library(tidyr)
library(ggplot2)
library(vroom)
library(purrr)
library(stringr)
library(sf)
library(knitr)
library(fixest)
```

# Link to the github project

<https://github.com/lounaduyck-afk/BARBARIT-DUYCK-Project.git>

# Loading and describing the data

## 1) Residential property market data

<https://www.data.gouv.fr/datasets/indicateurs-immobiliers-par-commune-et-par-annee-prix-et-volumes-sur-la-periode-2014-2024>

From this website, we downloaded 11 csv files, each file corresponding to a specific year from 2014 to 2024. Each file gathers aggregate variables related to residential property market at the municipality scale.

The data have been collected by the government, from 2014 to 2024. The latest update was on the 7th of July 2025.

### Methodology

Data come from the treatment of another datagouv database regarding the Request for Property Values. From this initial database, they took only single-sale transfers, for which the values lied between 15,000€ and 10,000,000€, the surface between 10m² and 250m² for apartments, and between 10m² and 400m² for houses.

The population is the residential property transactions in Mainland France. The transactions have been aggregated at the municipalities level, which means that each line corresponds to a municipality for a unique year. Each municipality is identified by its INSEE code.

It is important to note that this database covers the whole of metropolitan France, with the exception of the departments of Bas-Rhin, Haut-Rhin, and Moselle, where the relevant data fall under the *livre foncier* (land registry) system.

One of the main strengths of this database is that it provides a common reference framework over a near-national geographical scope, making it possible to compare territories with one another, to understand their specific characteristics, and to produce studies and indicators based on homogeneous methods.

### Data merging

```{r, warning=FALSE, message=FALSE, echo=FALSE}
annees <- 2014:2024

DVF_panel <- map_dfr(annees, function(a) {

  df <- vroom(here("Data", paste0("dvf", a, ".csv")))

  if (a <= 2023) {
    df <- df |>
      rename(
        PropMaison = propmaison,
        PropAppart = propappart
      )
  } else if (a == 2024) {
    df <- df |>
      rename(
        Annee = annee,
        Nb_mutations = nb_mutations
      )
  }

  df
}) |>
  select(-1) |>
  mutate(Annee = as.integer(Annee))

```

Explanations: after standardizing the labels, we merge our data using bind_rows in order to get panel data.

### Data descriptions

#### Basic Information

```{r, warning=FALSE, message=FALSE, echo=FALSE}
nb_columns <- ncol(DVF_panel)

nb_rows <- nrow(DVF_panel)

obs_by_year <- DVF_panel |>
  group_by(Annee) |>
  summarise(n_obs = n()) |>
  arrange(Annee)

summary_table <- tibble(
  Type = c("Columns", "Total Rows", paste("Observations", obs_by_year$Annee, sep=" ")),
  Number = c(nb_columns, nb_rows, obs_by_year$n_obs)
)

summary_table |>
  knitr::kable()
```

#### Variables descriptions

-   INSEE code: a qualitative variable which is unique for each municipality
-   Year
-   The number of transfers: quantitative variable which is the sum of the house sales and flat sales
-   The number of house sales: quantitative variable
-   The number of flat sales: quantitative variable
-   The proportion of house sales in percentage: quantitative variable
-   The proportion of flat sales in percentage: quantitative variable
-   The average price of properties sold in €: quantitative variable
-   The average price per square metre of properties sold in €: quantitative variable
-   The average surface area of properties sold in square metre: quantitative variable

## 2) Population evolution and stucture by municipalities

<https://www.insee.fr/fr/statistiques/8581696?sommaire=8581933>

From this website, we downloaded csv file corresponding to "Évolution et structure de la population en 2022 - Commune - France hors Mayotte"

We have population statistics in 2011, 2016 and 2022 for each municipality in France, except Mayotte, corresponding to the January 2025 geography.

They have been collected by the INSEE during the census.

The data cover population distribution by sex, age, occupations but also housing conditions and relocation.

```{r, warning=FALSE, message=FALSE, echo=FALSE}
Population <- vroom(here("Data","base-cc-evol-struct-pop-2022.csv"))
Labels <- vroom(here("Data","meta_base-cc-evol-struct-pop-2022.csv"))
```

### Data cleaning

#### Standardizing the labels

```{r, warning=FALSE, message=FALSE, echo=FALSE}
new_names <- setNames(Labels$COD_VAR, Labels$LIB_VAR)
Population <- Population |>
  rename(!!!new_names)
Population <- Population |>
  rename(INSEE_COM = `Arrondissement municipal`, "Pop 15 ans ou plus Autres personnes sans activité professionnelle en 2016 (compl)" = `Pop 15 ans ou plus Autres en 2016 (compl)`, "Pop 15 ans ou plus Hommes Autres personnes sans activité professionnelle en 2016 (compl)" = `Pop 15 ans ou plus Hommes Autres en 2016 (compl)`, "Pop 15 ans ou plus Femmes Autres personnes sans activité professionnelle en 2016 (compl)
" = `Pop 15 ans ou plus Femmes Autres en 2016 (compl)`, "Pop 15-24 ans Autres personnes sans activité professionnelle en 2016 (compl)
" = `Pop 15-24 ans Autres en 2016 (compl)`, "Pop 25-54 ans Autres personnes sans activité professionnelle en 2016 (compl)
" = `Pop 25-54 ans Autres en 2016 (compl)`, "Pop 55 ans ou plus Autres personnes sans activité professionnelle en 2016 (compl)" = `Pop 55 ans ou plus Autres en 2016 (compl)`)
```

Explanations: When we downloaded the data sets, we had two csv files, one with the data we want to look at, in which the labels had a specific code, and the other with the labels' code and their meaning. Then, we changed the code labels of our data set of interest to get their meaning.

### Modificating the structure of the dataset

```{r, warning=FALSE, message=FALSE, echo=FALSE}
Population <- Population |>
  pivot_longer(
    cols = -INSEE_COM,
    names_to = "temp",
    values_to = "value"
  ) |>
  mutate(
    Annee = str_extract(temp, "\\d{4}"),                
    Variable = str_remove(temp, " en \\d{4}.*$")               
  ) |>
  select(INSEE_COM, Annee, Variable, value) |>
  pivot_wider(
    names_from = Variable,
    values_from = value
  ) |>
  arrange(INSEE_COM, Annee)
```

Explanations:

Initially, we had one row for each municipality and the variables for each year. For example, we have 3 columns related to the number of men in the municipality, one for 2022, 2016 and 2011 ("Pop Homme en 2022", "Pop Homme en 2016", "Pop Homme en 2011"). Therefore, in order to get a panel data structure as the DVF data set we uploaded in the first part, we used at first pivot longer and we extract the year in each initial label. Then we create an additional column "Annee" and we used pivot wider to have the panel structure.

### Changing the columns type

```{r, echo=FALSE}
Population <- Population |>
  mutate(Annee = as.integer(Annee))
```

### Data descriptions

#### Basic Information

```{r, warning=FALSE, message=FALSE, echo=FALSE}
nb_columns <- ncol(Population)

nb_rows <- nrow(Population)

obs_by_year <- Population |>
  group_by(Annee) |>
  summarise(n_obs = n()) |>
  arrange(Annee)

summary_table <- tibble(
  Type = c("Columns", "Total Rows", paste("Observations", obs_by_year$Annee, sep=" ")),
  Number = c(nb_columns, nb_rows, obs_by_year$n_obs)
)

summary_table |>
  knitr::kable()
```

#### Variables description

-   Population: quantitative variable - number of residents in each municipality
-   Per age range population (0-14, 15-29, 30-44, 45-59, 60-74, 75-89, 90+)
-   Men population
-   Per age range men population (0-14, 15-29, 30-44, 45-59, 60-74, 75-89, 90+)
-   Women population
-   Per age range women population (0-14, 15-29, 30-44, 45-59, 60-74, 75-89, 90+)
-   Number of residents who live in the municipality for more than 1 year and who live in the same accommodation than 1 year ago
-   Number of residents who live in the municipality for more than 1 year and who live in a different accommodation than 1 year ago
-   Number of residents who lived outside the municipality 1 year ago (in a different department, in a different region, in Mainland France or in DOM) and their distribution by age and by sex.
-   Number of residents in each CSP for every age range and sex.

## 3) Unemployment

<https://www.observatoire-des-territoires.gouv.fr/taux-de-chomage-des-15-64-ans-rp>

From this website, we downloaded the file corresponding to "Taux de chômage des 15-64 ans", which was initially in the xlsx format. We therefore opened it on Excel and stored it as a csv file.

The data have been collected by the Insee, using the census in 2011, 2012, 2016 and 2022.

The data set gather, for each municipality and for each year of census, the total number of unemployed residents, the female unemployed residents and the male unemployed residents.

```{r, warning=FALSE, message=FALSE, echo=FALSE}
Unemployment <- vroom(here("Data","indic_sex_rp.csv"),
  delim = ";",
  skip = 4,
  col_names = TRUE,
  col_types = cols(
    codgeo = col_character(),
    libgeo = col_character(),
    an = col_integer(),
    sexe = col_factor(),
    nb_chom1564 = col_integer())) |>
  rename(INSEE_COM = codgeo,
         Annee = an,
         Chomeur = nb_chom1564)
```

```{r, echo=FALSE}
Unemployment |>
  filter(Annee == 2012) |>
  summarise(na_chomeur = sum(is.na(Chomeur)),
    obs = n()) |>
  knitr::kable()
```

Since all the observations in 2012 are NA, we apply a filter to remove them.

```{r, echo=FALSE}
Unemployment <- Unemployment |>
  filter(!Annee==2012)
```

### Modificating the structure of the dataset

```{r, echo=FALSE}
Unemployment <- Unemployment |>
  mutate(sexe = case_when(
    sexe == "F" ~ "chomeur_femme",
    sexe == "H" ~ "chomeur_homme",
    sexe == "T" ~ "chomeur_total"
  )) |>
  pivot_wider(
    names_from = sexe,
    values_from = Chomeur
  )
```

Initially, for each year, each municipality had three rows:

\- Total number of unemployed residents

\- The number of female unemployed residents

\- The number of male unemployed residents To get a consistent panel structure, we created additional columns, that specify to what each number refers, and we used pivot wider.

### Data description

#### Basic information

```{r, warning=FALSE, message=FALSE, echo=FALSE}
nb_columns <- ncol(Unemployment)

nb_rows <- nrow(Unemployment)

obs_by_year <- Unemployment |>
  group_by(Annee) |>
  summarise(n_obs = n()) |>
  arrange(Annee)

summary_table <- tibble(
  Type = c("Columns", "Total Rows", paste("Observations", obs_by_year$Annee, sep=" ")),
  Number = c(nb_columns, nb_rows, obs_by_year$n_obs)
)

summary_table |>
  knitr::kable()
```

#### Variables description

-   INSEE_COM: as the previous data bases, this one uses the Insee code to classify the municipality
-   libgeo: it corresponds to the name of the municipality associated with the INSEE code
-   Annee
-   Sex: coded as a factor, it expresses the gender category of the observation, distinguishing between females, males, and the total population
-   Chomeur: number of unemployed people (based on the INSEE's definition of unemployment)

# Merging the databases

## Code

```{r, warning=FALSE, message=FALSE, echo=FALSE}
DVFcensus <- DVF_panel |>
  filter(Annee %in% c(2022, 2016))

Population_filtered <- Population |>
  filter(Annee %in% c(2022, 2016))

Unemployment_filtered <- Unemployment |>
  filter(Annee%in%c(2016,2022))
```

```{r, warning=FALSE, message=FALSE, echo=FALSE}
Merge <- left_join(
  DVFcensus,
  Population_filtered,
  by = join_by(INSEE_COM, Annee)
)

Merge <- Merge |>
  left_join(
    Unemployment_filtered,
    by = join_by(INSEE_COM, Annee))
```

## Description

From the 3 data bases, we only kept data from 2016 and 2022, years of census and common to all bases. Using left join, because we wanted to keep all the observations we have in the initial panel set and match them with additional data we have from the two other data sets, we merged the data sets. Before merging, we ensured that the structure of the three data sets where coherent to be merged. We did this work in previous sections.

## Geographical data

​​<https://geoservices.ign.fr/adminexpress>

The ADMIN EXPRESS product line centralizes five related datasets: ADMIN EXPRESS, ADMIN EXPRESS COG, ADMIN EXPRESS COG CARTO, ADMIN EXPRESS COG CARTO Petite Échelle, and ADMIN EXPRESS COG CARTOPLUS PE. Together, these datasets provide detailed administrative boundary layers for France across multiple levels. They include the following geographic units:

-   Arrondissements (departmental and municipal)

-   Cantons

-   Administrative seats (chef-lieux) at various levels: arrondissement, municipal arrondissement, EPCI, canton, territorial collectivity, commune, associated or delegated commune, département, and region

-   Communes and associated or delegated communes

-   Départements

-   EPCIs (public inter-municipal cooperation bodies)

-   Regions

The main product, ADMIN EXPRESS, provides large-scale administrative boundaries and is updated monthly. An additional edition, ADMIN EXPRESS COG, is aligned with the Code Officiel Géographique (COG) published annually by INSEE.

This product line facilitates combining administrative boundaries with external datasets to build thematic representations of the French territory at different administrative levels. For our project, we mainly rely on the commune layer, as it gives the spatial unit necessary to merge geographic boundaries with demographic, employment, and real-estate datasets at the municipal level.

Here, we import both layers using `st_read()`, then standardize their INSEE municipality codes by trimming extra spaces and ensuring they always have five digits. This harmonization is essential for correctly joining the geographic layers with our DVF database, which also uses INSEE codes as identifiers.

```{r, warning=FALSE, message=FALSE, echo=FALSE}
gpkg_path <- "Data/ADE_4-0_GPKG_LAMB93_FXX-ED2025-10-15.gpkg"
communes_sf <- st_read(gpkg_path, layer = "chef_lieu_de_commune") |>
mutate(code_insee_de_la_commune = sprintf("%05s", trimws(code_insee_de_la_commune)))

communes_poly <- st_read(gpkg_path, layer = "commune") |>
mutate(code_insee = sprintf("%05s", trimws(code_insee)))
```

## 4) Department labels data

Source: https://www.data.gouv.fr/datasets/departements-de-france

This file gather: - The department code - The department name - The region code - The region name

We use it here to add to our Merge data set the name of the department to which each municipality belongs. This will be useful latter to do some data visualization at the department level.

```{r, echo=FALSE}
dep <- vroom(here("Data","departements-france.csv"))
```

We extract from the "INSEE code commune" the numbers linked to the department code.

```{r, echo=FALSE}
Merge <- Merge |>
  mutate(
    code_departement = substr(as.character(INSEE_COM), 1, 2)
  ) |>
  left_join(
    dep |> select(code_departement, nom_departement),
    by = "code_departement"
  )
```

```{r, echo=FALSE}
sum(is.na(Merge$code_departement))
```

Only 2 municipalities do not have a departement code. This is due to two NA for INSEE_COM, one for 2016, the other for 2022.

# Descriptive analysis of the structure of the French real estate market

```{r, warning=FALSE, message=FALSE, echo=FALSE}

df_both <- DVF_panel |>
  filter(Annee %in% c(2016, 2022)) |>
  select(INSEE_COM, Annee, Prixm2Moyen) |>
  mutate(Prixm2Moyen = ifelse(Prixm2Moyen <= 0, NA, Prixm2Moyen))

carte_poly_both <- bind_rows(
  communes_poly |> mutate(Annee = 2016),
  communes_poly |> mutate(Annee = 2022)
) |>
  left_join(
    df_both,
    by = c("code_insee" = "INSEE_COM", "Annee" = "Annee")
  ) |>
  mutate(Annee = factor(Annee))

ggplot(carte_poly_both) +
  geom_sf(aes(fill = Prixm2Moyen), color = NA) +
  scale_fill_viridis_c(
    trans = "log", direction = -1,
    na.value = "grey70"
  ) +
  facet_wrap(~ Annee) +
  labs(
    title = "Average Price per square meter per Municipality",
    fill  = "Average Price per m² (€)"
  ) +
  theme_minimal()

```

Between 2016 and 2022, the maps reveal an almost global increase in average prices per square meter, particularly marked in and around large metropolitan areas (we can distinguish Paris, Lyon, Marseille, Bordeaux, Toulouse...). This trend is followed by a significant gap between urban and rural areas, reflecting the dynamics of metropolitanization and differentiated attractiveness. The spread of high price levels in cities such as Bordeaux, Annecy, and Nantes suggests that the rise in prices extends beyond the Paris region and reflects common macroeconomic determinants combined with heterogeneous local trends. Coastal municipalities also have high prices per square meter, which can be explained by several factors such as tourist appeal, quality of life, and the presence of many second homes. From an economic point of view, these maps already allow us to make the following assumptions: prices per square meter seem to be correlated with several factors, including urbanity, unemployment, and the attractiveness of different areas.\
These maps allow us to easily distinguish between attractive areas of France, such as cities and coastal regions, and less attractive areas, known as the “diagonale du vide” (empty diagonal), which stretches from the northeast to the west of the country.

### Ile de France map

```{r, warning=FALSE, message=FALSE, echo=FALSE}
departements_idf <- c("75","92","93","94","91","77","78","95")

communes_idf <- communes_poly |>
mutate(code_insee = sprintf("%05s", trimws(code_insee))) |>
filter(substr(code_insee, 1, 2) %in% departements_idf)

carte_idf_2016 <- communes_idf |>
  left_join(DVF_panel |> filter(Annee == 2016),
            by = c("code_insee" = "INSEE_COM")) |>
  mutate(Annee = "2016")

carte_idf_2022 <- communes_idf |>
  left_join(DVF_panel |> filter(Annee == 2022),
            by = c("code_insee" = "INSEE_COM")) |>
  mutate(Annee = "2022")

carte_idf_1622 <- bind_rows(carte_idf_2016, carte_idf_2022)

# Graph with facet_wrap
ggplot(carte_idf_1622) +
  geom_sf(aes(fill = Prixm2Moyen), color = "white", size = 0.1) +
  scale_fill_viridis_c(trans = "log", direction = -1) +
  theme_minimal() +
  labs(
    title = "Average price per square meter — Île-de-France",
    fill  = "Average Price per square meter (€)"
  ) +
  facet_wrap(~ Annee)

```

### Graphic of the evolution of the average price of real estate per square meter

```{r, warning=FALSE, message=FALSE, echo=FALSE}
DVF_panel |>
  group_by(Annee) |>
  summarise(prix_moyen_m2 = mean(Prixm2Moyen, na.rm = TRUE)) |>
  ggplot(aes(x = Annee, y = prix_moyen_m2)) +
  geom_line(linewidth = 1.2, color = "steelblue") +
  geom_point(size = 2, color = "steelblue") +
  theme_minimal() +
    labs(
    title = "Evolution of the Average Property Price per square meter",
    x = "Year",
    y = "Average price per m² (€)"
  )
```

This graph confirms the trend observed on the maps: average real estate prices per square meter rose sharply in France between the mid-2010s and the mid-2020s. In fact, prices remained fairly stable between 2014 and 2019, hovering around 1600€ per m², with a slight decline observed in 2018.

Between 2019 and 2023, the average price per square meter in France rose sharply from 1600€ to almost 2000€.\
This trend probably reflects several factors: very favorable borrowing conditions for most of the period due to low interest rates, stronger demand for real estate after the Covid-19 crisis, and a possible move to more attractive regions.

A downward trend is observed since 2023. This recent slowdown could be linked to the change in monetary policy and rising interest rates, which have reduced households' purchasing power in the real estate market. Overall, this graph highlights a change in the market regime, which has shifted from a relatively stable situation at the outset to a phase of significant price appreciation that appears to be highly sensitive to macroeconomic and financial shocks.

### Graphic of the evolution of the number of transactions

```{r, warning=FALSE, message=FALSE, echo=FALSE}
DVF_panel |>
  group_by(Annee) |>
  summarise(Total_mutations = sum(Nb_mutations, na.rm = TRUE)) |>
  ggplot(aes(x = Annee, y = Total_mutations)) +
  geom_line(color = "steelblue", size = 1.2) +
  geom_point(color = "red", size = 2) +
  labs(title = "Evolution of the number of real estate transactions in France (2014-2024)",
       x = "Year",
       y = "Number of transactions") +
  theme_minimal()
```

In this graphical representation we observe that the volume of transactions increased between 2014 and 2016, suggesting a gradual recovery in the real estate market after the sovereign debt crisis in the eurozone (which peaked in 2010-2012).\
A sharp decline was observed in 2017-2018, which could reflect regulatory changes, data coverage issues, or tighter credit conditions that temporarily reduced market activity.

From 2019, the number of transactions increased strongly, exceeding one million in 2021 and remaining at a high level in 2022. This expansion is consistent with the exceptionally low interest rates observed before 2022 and with the sharp increase in demand for housing around the Covid-19 crisis, when households sought larger or better-located homes.

After 2022, volumes decline sharply until 2023-2024, indicating a slowdown in the market. This slowdown shows rising mortgage rates and inflation, which have eroded purchasing power and increased the cost of financing.

### Graphical representation of the Average property size over time

```{r, warning=FALSE, message=FALSE, echo=FALSE}
DVF_panel |>
  group_by(Annee) |>
  summarise(SurfaceMoy = mean(SurfaceMoy, na.rm = TRUE)) |>
  ggplot(aes(x = Annee, y = SurfaceMoy)) +
  geom_line(color = "darkred", size = 1.2) +
  geom_point(color = "red", size = 2) +
  labs(title = "Evolution of the average surface area of properties in France",
       x = "Year",
       y = "Average surface area (m²)") +
  theme_minimal()
```

We see that there was a gradual increase in the average surface area of propeties between 2014 and 2016, with the average size rising from just under 100 m² to around 102 m². This upward trend continued moderately until 2020, suggesting that buyers gradually turned to larger homes, perhaps in response to changing preferences and improved living standards. A peak is observed in 2021, above 103 m², which could reflect the impact of the Covid-19 crisis: lockdowns and teleworking have increased the value of space, encouraging households to look for homes with additional rooms, gardens, or offices.

After 2021, the average surface area decreases, stabilizing at around 100 m² in 2023. This reversal could be linked to rising property prices and mortgage rates, which have stretched budgets and forced households to buy smaller homes.

### Distribution of houses vs. apartments by year

```{r, warning=FALSE, message=FALSE, echo=FALSE}

# 1) Unique aggregation
df_agrege <- DVF_panel |>
  group_by(Annee) |>
  summarise(
    Maisons      = sum(NbMaisons, na.rm = TRUE),
    Appartements = sum(NbApparts, na.rm = TRUE),
    Total_mutations = sum(Nb_mutations, na.rm = TRUE),   # <- ton layer courbe
    .groups = "drop"
  ) |>
  mutate(Annee = as.numeric(Annee))

# 2. Long format for the barres
df_long <- df_agrege |>
  select(Annee, Maisons, Appartements) |>
  pivot_longer(
    cols = c(Maisons, Appartements),
    names_to  = "TypeBien",
    values_to = "Nombre"
  )

# 3. Graph
ggplot() +
  geom_col(
    data = df_long,
    aes(x = Annee, y = Nombre, fill = TypeBien),
    position = "stack",
    alpha = 0.85
  ) +
  geom_line(
    data = df_agrege,
    aes(x = Annee, y = Total_mutations),
    size = 1.1
  ) +
  geom_point(
    data = df_agrege,
    aes(x = Annee, y = Total_mutations),
    size = 2
  ) +
  labs(
    title = "Distribution of houses and apartments with total transactions",
    x = "Year",
    y = "Number of transactions",
    fill = "Property type"
  ) +
  theme_minimal()

```

```{r, warning=FALSE, message=FALSE, echo=FALSE}
DVF_panel |>
  group_by(Annee) |>
  summarise(
    TotalMaisons = sum(NbMaisons, na.rm = TRUE),
    TotalApparts = sum(NbApparts, na.rm = TRUE)
  ) |>
  mutate(
    Total = TotalMaisons + TotalApparts,
    Houses = TotalMaisons / Total,
    Appartments = TotalApparts / Total
  ) |>
  pivot_longer(
    cols = c(Houses, Appartments),
    names_to = "TypeBien",
    values_to = "Share"
  ) |>
  ggplot(aes(x = Annee, y = Share, fill = TypeBien)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Distribution of houses and apartments per year",
    x = "Year",
    y = "Share of transactions (%)"
  ) +
  theme_minimal()

```

In the first graphical representation we observe that the volume of transactions increased between 2014 and 2016, suggesting a gradual recovery in the real estate market after the sovereign debt crisis in the eurozone (which peaked in 2010-2012). A significant decline was observed in 2017-2018, which could reflect regulatory changes, data coverage issues, or tighter credit conditions that temporarily reduced market activity.

From 2019, the number of transactions increased strongly, exceeding one million in 2021 and remaining at a high level in 2022. This expansion is consistent with the exceptionally low interest rates observed before 2022 and with the sharp increase in demand for housing around the Covid-19 crisis, when households sought larger or better-located homes.

After 2022, volumes decline sharply until 2023-2024, indicating a slowdown in the market. This slowdown reflects rising mortgage rates and inflation, which have eroded purchasing power and increased the cost of financing.

However, the composition chart shows that, despite these pronounced fluctuations in activity, the market structure has remained remarkably stable: apartments have consistently accounted for just over half of transactions, around 55-58%, and houses around 42-45%. Only minor variations can be observed from year to year, with no lasting shift in favor of one type of housing over another.

## Comparison across the 10 largest French metropolitan areas

```{r, warning=FALSE, message=FALSE, echo=FALSE}
metro_labels <- tibble::tibble(
  INSEE_COM = c("75056", "69123", "33063", "13055", "31555",
                "06088", "44109", "34172", "67482", "59350"),
  Metro = c("Paris",
            "Lyon",
            "Bordeaux",
            "Marseille",
            "Toulouse",
            "Nice",
            "Nantes",
            "Montpellier",
            "Strasbourg",
            "Lille")
)
```

```{r, warning=FALSE, message=FALSE, echo=FALSE}
df_comp <- DVF_panel |>
  filter(INSEE_COM %in% metro_labels$INSEE_COM) |>
  left_join(metro_labels, by = "INSEE_COM")
```

```{r, warning=FALSE, message=FALSE, echo=FALSE}
ggplot(df_comp, aes(x = Annee, y = Prixm2Moyen, color = Metro)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2) +
  theme_minimal() +
  labs(
    title = "Comparison across the 10 largest French metropolitan areas (m²)",
    x = "Year",
    y = "Average property price (€)",
    color = "Metropolitan area"
  ) +
  scale_color_brewer(palette = "Set2")
```

```{r, warning=FALSE, message=FALSE, echo=FALSE}
df_comp |>
  filter(!is.na(Nb_mutations)) |>
  ggplot(aes(x = Annee, y = Nb_mutations, color = Metro)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.8) +
  theme_minimal() +
  labs(
    title = "Comparison of transaction volumes across the 10 French metropolitan areas",
    x = "Year",
    y = "Number of transactions",
    color = "Metropolitan area"
  ) +
scale_color_brewer(palette = "Set2")
```

The first graph highlights an upward trend in real estate prices in the ten largest French cities between 2016 and 2022, confirming the previous observations. Paris clearly stands out as an outlier, with prices almost twice as high as those in other cities, reflecting very limited land supply and sustained investor demand. Most regional cities such as Lyon and Nice are also experiencing significant increases, suggesting growing appeal and demographic pressure in major urban centers.

In the second graph we observe that the sharp decline in Paris around 2017, followed by a rebound after 2019, reflects sensitivity to credit conditions and affordability constraints. Regional cities show approximatively the same shape, but in smaller proportions, indicating more elastic supply and a less stained real estate market outside the capital.

Taken together, these figures suggest that the French real estate market is becoming increasingly polarized: high prices in Paris may crowd out some demand, while regional metropolitan areas continue to grow in both value and activity. This requires a differentiated regional analysis when evaluating housing or territorial policies.

## Comparison between departments

```{r, warning=FALSE, message=FALSE, echo=FALSE}
DVF_panel |>
  mutate(dep = substr(INSEE_COM, 1, 2)) |>
  filter(
    !dep %in% c("97", "98", "57", "68"), 
    dep != "2A", dep != "2B",
    !is.na(Prixm2Moyen)
  ) |>
  group_by(dep, Annee) |>
  summarise(
    prix_m2 = mean(Prixm2Moyen, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(x = Annee, y = prix_m2)) +
  geom_line(linewidth = 0.6, color = "steelblue") +
  geom_point(size = 1.2, color = "steelblue") +
  facet_wrap(~ dep, scales = "free_y") +  
  theme_minimal() +
  labs(
    title = "Evolution of average price per square meter by department",
    x = "Year",
    y = "Average price per m² (€)"
  )
```

```{r, echo=FALSE}
# =========================
# 2022 — Price per m² (Departments)
# =========================
Merge |>
  filter(Annee == 2022, !is.na(code_departement), !is.na(Prixm2Moyen)) |>
  group_by(code_departement, nom_departement) |>
  summarise(
    Prixm2Moyen = mean(Prixm2Moyen, na.rm = TRUE),
    n_communes  = sum(!is.na(Prixm2Moyen)),
    .groups = "drop"
  ) |>
  filter(is.finite(Prixm2Moyen), n_communes > 0) |>
  arrange(Prixm2Moyen) -> p_2022

p_2022 |>
  slice_head(n = 10) |>
  transmute(
    `Lowest prices (2022)` = nom_departement,
    `Avg price (€/m²)`     = round(Prixm2Moyen, 0)
  ) -> low_2022

p_2022 |>
  slice_tail(n = 10) |>
  arrange(desc(Prixm2Moyen)) |>
  transmute(
    `Highest prices (2022)` = nom_departement,
    `Avg price (€/m²)`      = round(Prixm2Moyen, 0)
  ) -> high_2022

bind_cols(low_2022, high_2022) |>
  kable(
    digits = 0,
    caption = "Housing prices — Departments with the lowest and highest average price per m² (2022)",
    align = "l"
  )


# =========================
# 2016 — Price per m² (Departments)
# =========================
Merge |>
  filter(Annee == 2016, !is.na(code_departement), !is.na(Prixm2Moyen)) |>
  group_by(code_departement, nom_departement) |>
  summarise(
    Prixm2Moyen = mean(Prixm2Moyen, na.rm = TRUE),
    n_communes  = sum(!is.na(Prixm2Moyen)),
    .groups = "drop"
  ) |>
  filter(is.finite(Prixm2Moyen), n_communes > 0) |>
  arrange(Prixm2Moyen) -> p_2016

p_2016 |>
  slice_head(n = 10) |>
  transmute(
    `Lowest prices (2016)` = nom_departement,
    `Avg price (€/m²)`     = round(Prixm2Moyen, 0)
  ) -> low_2016

p_2016 |>
  slice_tail(n = 10) |>
  arrange(desc(Prixm2Moyen)) |>
  transmute(
    `Highest prices (2016)` = nom_departement,
    `Avg price (€/m²)`      = round(Prixm2Moyen, 0)
  ) -> high_2016

bind_cols(low_2016, high_2016) |>
  kable(
    digits = 0,
    caption = "Housing prices — Departments with the lowest and highest average price per m² (2016)",
    align = "l"
  )
```

```{r, warning=FALSE, message=FALSE, echo=FALSE}
DVF_panel |>
  mutate(dep = substr(INSEE_COM, 1, 2)) |>
  filter(
    !dep %in% c("97", "98", "57", "68"),  
    dep != "2A", dep != "2B",
    !is.na(SurfaceMoy)
  ) |>
  group_by(dep, Annee) |>
  summarise(
    surface_moy = mean(SurfaceMoy, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(x = Annee, y = surface_moy)) +
  geom_line(linewidth = 0.6, color = "steelblue") +
  geom_point(size = 1.2, color = "steelblue") +
  facet_wrap(~ dep) +   
  theme_minimal() +
  labs(
    title = "Evolution of average surface area by department",
    x = "Year",
    y = "Average surface area (sqm)"
  )

```

```{r, echo=FALSE}
# =========================
# 2022 — Number of transactions (mutations)
# =========================
Merge |>
  filter(Annee == 2022, !is.na(code_departement), !is.na(Nb_mutations)) |>
  group_by(code_departement, nom_departement) |>
  summarise(
    Nb_mutations = sum(Nb_mutations, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(is.finite(Nb_mutations), Nb_mutations > 0) |>
  arrange(Nb_mutations) -> m_2022

m_2022 |>
  slice_head(n = 10) |>
  transmute(
    `Lowest mutations (2022)` = nom_departement,
    `Mutations`               = round(Nb_mutations, 0)
  ) -> low_2022

m_2022 |>
  slice_tail(n = 10) |>
  arrange(desc(Nb_mutations)) |>
  transmute(
    `Highest mutations (2022)` = nom_departement,
    `Mutations`                = round(Nb_mutations, 0)
  ) -> high_2022

bind_cols(low_2022, high_2022) |>
  kable(
    digits = 0,
    caption = "Real-estate transactions — Departments with the lowest and highest number of mutations (2022)",
    align = "l"
  )


# =========================
# 2016 — Number of transactions (mutations)
# =========================
Merge |>
  filter(Annee == 2016, !is.na(code_departement), !is.na(Nb_mutations)) |>
  group_by(code_departement, nom_departement) |>
  summarise(
    Nb_mutations = sum(Nb_mutations, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(is.finite(Nb_mutations), Nb_mutations > 0) |>
  arrange(Nb_mutations) -> m_2016

m_2016 |>
  slice_head(n = 10) |>
  transmute(
    `Lowest mutations (2016)` = nom_departement,
    `Mutations`               = round(Nb_mutations, 0)
  ) -> low_2016

m_2016 |>
  slice_tail(n = 10) |>
  arrange(desc(Nb_mutations)) |>
  transmute(
    `Highest mutations (2016)` = nom_departement,
    `Mutations`                = round(Nb_mutations, 0)
  ) -> high_2016

bind_cols(low_2016, high_2016) |>
  kable(
    digits = 0,
    caption = "Real-estate transactions — Departments with the lowest and highest number of mutations (2016)",
    align = "l"
  )
```

## Exploring joint distributions
### Joint distribution of average price per m² and average surface area by Department
```{r, warning=FALSE, message=FALSE, echo=FALSE}

DVF_panel |>
  mutate(dep = substr(INSEE_COM, 1, 2)) |>
  filter(
    !dep %in% c("97", "98", "57", "68"),
    dep != "2A", dep != "2B",
    !is.na(Prixm2Moyen),
    !is.na(SurfaceMoy)
  ) |>
  group_by(dep, Annee) |>
  summarise(
    prix_m2 = mean(Prixm2Moyen, na.rm = TRUE),
    surface_moy = mean(SurfaceMoy, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(x = surface_moy, y = prix_m2)) +
  geom_point(alpha = 0.7, size = 2, color = "steelblue") +
  theme_minimal() +
  labs(
    title = "Joint distribution of average price per m² and average surface area",
    x = "Average surface area (sqm)",
    y = "Average price per m² (€)"
  ) +
  geom_smooth(method = "lm", se = FALSE, color = "darkred")

```

In this graph, each point represents a department at one precise year. The scatter plot shows a negative relationship, which means that transactions involving larger homes tend to be associated with lower prices per square meter, while smaller properties command higher unit values. The dense scatter plot between approximately 80 and 110 m² shows that most French sales fall within this range, with prices largely clustered around 1500€ to 3000€ per m².

The extreme observations in the upper left correspond to very high prices for relatively small areas, probably reflecting Paris and a few particularly sought-after urban areas where land scarcity and high demand drive up valuations. This trend shows that price heterogeneity is determined not only by location, but also by the composition of the housing stock. Areas with a high proportion of small apartments have higher prices per square meter, while areas where houses and spacious dwellings are the norm have lower unit prices. These data confirm the need to control for average size or type of dwelling in any econometric analysis of real estate prices.

### Joint distribution of average price per m² and average surface area by Communes
```{r}
DVF_panel |>
  mutate(dep = substr(INSEE_COM, 1, 2)) |>
  filter(
    !dep %in% c("97", "98", "57", "68"),
    dep != "2A", dep != "2B",
    !is.na(INSEE_COM),
    !is.na(Prixm2Moyen),
    !is.na(SurfaceMoy)
  ) |>
  group_by(INSEE_COM, Annee) |>
  summarise(
    prix_m2     = mean(Prixm2Moyen, na.rm = TRUE),
    surface_moy = mean(SurfaceMoy,  na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(x = surface_moy, y = prix_m2)) +
  geom_point(alpha = 0.08, size = 0.8, color = "steelblue") +
  theme_minimal() +
  labs(
    title = "Joint distribution of average price per m² and average surface area (municipality level)",
    x = "Average surface area (sqm)",
    y = "Average price per m² (€)"
  ) +
  geom_smooth(method = "lm", se = FALSE, color = "darkred")

```


### Joint distribution of average price per m² and share of apartments by Departments
```{r, warning=FALSE, message=FALSE, echo=FALSE}
DVF_panel |>
  mutate(dep = substr(INSEE_COM, 1, 2)) |>
  filter(
    !dep %in% c("97", "98", "57", "68"),
    dep != "2A", dep != "2B",
    !is.na(Prixm2Moyen),
    !is.na(NbApparts),
    !is.na(Nb_mutations)
  ) |>
  group_by(dep, Annee) |>
  summarise(
    prix_m2 = mean(Prixm2Moyen, na.rm = TRUE),
    part_apparts = sum(NbApparts, na.rm = TRUE) / sum(Nb_mutations, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(x = part_apparts, y = prix_m2)) +
  geom_point(alpha = 0.7, size = 2, color = "steelblue") +
  geom_smooth(method = "lm", se = FALSE, color = "darkred") +
  theme_minimal() +
  scale_x_continuous(labels = scales::percent_format()) +
  labs(
    title = "Joint distribution of average price per m² and share of apartments",
    subtitle = "Apartment share as a proxy for urban density",
    x = "Share of apartments (%)",
    y = "Average price per m² (€)"
  )

```

In this graph, the proportion of apartments is used as a proxy of urban density (a high propotion of apartments is the sign of a urbanized area). The scatter plot reveals a positive and approximately linear relationship, showing that municipalities with a higher proportion of apartments tend to have higher prices per m², while areas dominated by low proportions of apartments have the lowest unit prices.

A large number of observations below 25% apartments show that most French municipalities remain relatively low in density, with prices generally below 2000€ per m². The points on the far right, close to 100%, correspond to very dense urban areas, most likely Paris, where land scarcity, amenities, and the concentration of skilled households generate significantly higher valuations.

The slope suggests that price heterogeneity is strongly linked to the structure of the housing stock: compact real estate markets benefit more from location advantages than spacious residential markets. Therefore, any econometric assessment of real estate prices should take into account the proportion of apartments or the composition of the housing stock in order to avoid attributing the effects of density to other variables such as unemployment or income alone.

### Joint distribution of average price per m² and share of apartments by Communes
```{r}
DVF_panel |>
  mutate(dep = substr(INSEE_COM, 1, 2)) |>
  filter(
    !dep %in% c("97", "98", "57", "68"),
    dep != "2A", dep != "2B",
    !is.na(INSEE_COM),
    !is.na(Prixm2Moyen),
    !is.na(NbApparts),
    !is.na(Nb_mutations),
    Nb_mutations > 0
  ) |>
  group_by(INSEE_COM, Annee) |>
  summarise(
    prix_m2 = mean(Prixm2Moyen, na.rm = TRUE),
    part_apparts = sum(NbApparts, na.rm = TRUE) / sum(Nb_mutations, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(x = part_apparts, y = prix_m2)) +
  geom_point(alpha = 0.08, size = 0.8, color = "steelblue") +
  geom_smooth(method = "lm", se = FALSE, color = "darkred") +
  theme_minimal() +
  scale_x_continuous(labels = scales::percent_format()) +
  labs(
    title = "Joint distribution of average price per m² and share of apartments (municipality level)",
    subtitle = "Apartment share as a proxy for urban density",
    x = "Share of apartments (%)",
    y = "Average price per m² (€)"
  )

```


### Joint distribution of average price per m² and number of transactions by Department
```{r, warning=FALSE, message=FALSE, echo=FALSE}

DVF_panel |>
  mutate(dep = substr(INSEE_COM, 1, 2)) |>
  filter(
    !dep %in% c("97", "98", "57", "68"),
    dep != "2A", dep != "2B",
    !is.na(Prixm2Moyen),
    !is.na(Nb_mutations)
  ) |>
  group_by(dep, Annee) |>
  summarise(
    prix_m2 = mean(Prixm2Moyen, na.rm = TRUE),
    nb_mut = sum(Nb_mutations, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(x = nb_mut, y = prix_m2)) +
  geom_point(alpha = 0.7, size = 2, color = "steelblue") +
  geom_smooth(method = "lm", se = FALSE, color = "darkred") +
  theme_minimal() +
  labs(
    title = "Joint distribution of average price per m² and number of transactions",
    subtitle = "Transaction volume as a proxy for market activity",
    x = "Number of transactions",
    y = "Average price per m² (€)"
  )
```

Here we can observe a positive relationship showing that, on average, municipalities with higher transaction volumes tend to have higher prices per square meter. A very dense concentration of points below 10000 transactions and 2000€ per m² shows that the majority of French municipalities correspond to small and relatively inexpensive local markets.

The observations between 15000 and 30000 transactions correspond to large metropolitan or suburban areas characterized by high demand and many amenities. However, the cloud becomes increasingly heterogeneous as prices rise. Several high-priced municipalities have limited transaction volumes, which means that expensive markets are not necessarily the most active.

In this graph, we see the coexistence of two mechanisms: in dynamic urban areas, high activity could reflect intense demand that drives up prices, while in expensive locations, affordability constraints and supply rigidity may limit the number of sales.

### Joint distribution of average price per m² and number of transactions by Commune
```{r}
DVF_panel |>
  mutate(dep = substr(INSEE_COM, 1, 2)) |>
  filter(
    !dep %in% c("97", "98", "57", "68"),
    dep != "2A", dep != "2B",
    !is.na(INSEE_COM),
    !is.na(Prixm2Moyen),
    !is.na(Nb_mutations),
    Nb_mutations > 0
  ) |>
  group_by(INSEE_COM, Annee) |>
  summarise(
    prix_m2 = mean(Prixm2Moyen, na.rm = TRUE),
    nb_mut  = sum(Nb_mutations, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(x = log(nb_mut), y = prix_m2)) +
  geom_point(alpha = 0.08, size = 0.8, color = "steelblue") +
  geom_smooth(method = "lm", se = FALSE, color = "darkred") +
  theme_minimal() +
  labs(
    title = "Joint distribution of average price per m² and number of transactions (municipality level)",
    subtitle = "Log number of transactions",
    x = "Log number of transactions",
    y = "Average price per m² (€)"
  )

```



## Demographic indicators

To go further in the data analysis process, we calculate several indicators.

-   **The upper socio-economic group rate**: percentage of managers and higher intellectual professions among the population of 15+ years old.

-   **The unemployment rate**: the percentage of unemployed people among the active population (between 20 years old and 64 years old).

    It should be noted that this ratio is approximate, as it does not take into account individuals aged 15 to 19. As we did not have the number of inhabitants for each age group in each municipality, we chose the age group closest to the working population.

-   **The relocation rate**: the percentage of people that initiate a transaction during the year while they were not resident in this municipality previously.

```{r, echo=FALSE}
Merge <- Merge |>   
  mutate(TauxCSPsup =`Pop 15 ans ou plus Cadres, Prof. intel. sup.`/ `Pop 15 ans ou plus`*100,
         TauxChom = chomeur_total /(`Pop Hommes 20-64 ans`+`Pop Femmes 20-64 ans`)*100,
         TauxDem = (1 - (`Pop 1 an ou plus habitant 1 an avt même logt`+`Pop 1 an ou plus habitant 1 an avt autre logt même commune`)/Population)*100)
```

### The upper socio-economic group rate representation

```{r, echo=FALSE}
#Merging the geolocalisation dataset with the demographic indicators for each municipality
map_demo <- communes_sf |>
  left_join(
    Merge |> filter(Annee %in% c(2016, 2022)),
    by = c("code_insee_de_la_commune" = "INSEE_COM")) |>
  select(code_insee_de_la_commune, Annee, TauxCSPsup, TauxChom, TauxDem)
```

```{r, echo=FALSE}
map_demo |>
  filter(!is.na(TauxCSPsup)) |>
  ggplot() +
  geom_sf(aes(color = TauxCSPsup), size = 1, alpha = 0.7) +
  facet_wrap(~ Annee, ncol = 2, scales = "fixed") +
  scale_color_viridis_c(direction = -1) +
  labs(
    title = "CSP+ rate by municipalities",
    color = "Rate (%)") +
  theme_minimal()
```

### The unemployment rate representation

```{r, echo=FALSE}
map_demo |>
  filter(!is.na(TauxChom)) |>
  ggplot() +
  geom_sf(aes(color = TauxChom), size = 1, alpha = 0.7) +
  facet_wrap(~ Annee, ncol = 2, scales = "fixed") +
  scale_color_viridis_c(direction = -1) +
  labs(
    title = "Unemployment rate by municipalities",
    color = "Rate (%)"
  ) +
  theme_minimal()
```

### The relocation rate representation

```{r, echo=FALSE}
map_demo |>
  filter(!is.na(TauxDem)) |>
  ggplot() +
  geom_sf(aes(color = TauxDem), size = 1, alpha = 0.7) +
  facet_wrap(~ Annee, ncol = 2, scales = "fixed") +
  scale_color_viridis_c(direction = -1) +
  labs(
    title = "Relocation rate by municipalities",
    color = "Rate (%)"
  ) +
  theme_minimal()
```

### Analysis per departments

#### Distribution of socio-professional groups across departments

```{r, echo=FALSE}
csp_cols <- c(
  "Pop 15 ans ou plus Agriculteurs exploitants",
  "Pop 15 ans ou plus Artisans, Comm., Chefs entr.",
  "Pop 15 ans ou plus Cadres, Prof. intel. sup.",
  "Pop 15 ans ou plus Prof. intermédiaires ",
  "Pop 15 ans ou plus Employés",
  "Pop 15 ans ou plus Ouvriers",
  "Pop 15 ans ou plus Retraités",
  "Pop 15 ans ou plus Autres personnes sans activité professionnelle")

Merge |>
  filter(Annee %in% c(2016, 2022)) |>
  
  filter(!is.na(code_departement)) |>
  mutate(
    dep_label = paste0(code_departement, " – ", nom_departement),
    Annee = factor(Annee, levels = c("2016","2022"))
  ) |>
  pivot_longer(
    cols = all_of(csp_cols),
    names_to = "variable",
    values_to = "pop"
  ) |>
  mutate(
    CSP = str_remove(variable, "^Pop 15 ans ou plus\\s*"),
    CSP = str_squish(CSP)
  ) |>
  group_by(dep_label, Annee, CSP) |>
  summarise(pop = sum(pop, na.rm = TRUE), .groups = "drop") |>
  group_by(dep_label, Annee) |>
  mutate(part = pop / sum(pop)) |>
  ggplot(aes(x = Annee, y = part, fill = CSP)) +
  geom_col(width = 0.8) +
  coord_flip() +
  facet_wrap(~ dep_label, ncol = 3) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Distribution of socio-professional groups across departments.",
    subtitle = "Comparison between 2016 and 2022 (population aged 15 and over)",
    x = NULL,
    y = "Share",
    fill = "Socio-professional groups"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold", size = 9),
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(size = 8),
    panel.grid.major.y = element_blank()
  )
```

```{r, echo=FALSE}
# =========================
# 2022
# =========================
Merge |>
  filter(Annee == 2022, !is.na(code_departement)) |>
  group_by(code_departement, nom_departement) |>
  summarise(
    cadres = sum(`Pop 15 ans ou plus Cadres, Prof. intel. sup.`, na.rm = TRUE),
    denom  = sum(`Pop 15 ans ou plus`, na.rm = TRUE),
    TauxCSPsup = 100 * cadres / denom,
    .groups = "drop"
  ) |>
  filter(is.finite(TauxCSPsup), denom > 0) |>
  arrange(TauxCSPsup) -> d_2022

d_2022 |>
  slice_head(n = 10) |>
  transmute(
    `Lowest shares (2022)` = nom_departement,
    `CSP+ share` = round(TauxCSPsup, 2)
  ) -> low_2022

d_2022 |>
  slice_tail(n = 10) |>
  arrange(desc(TauxCSPsup)) |>
  transmute(
    `Highest shares (2022)` = nom_departement,
    `CSP+ share` = round(TauxCSPsup, 2)
  ) -> high_2022

bind_cols(low_2022, high_2022) |>
  kable(
    digits = 2,
    caption = "CSP+ — Departments with the lowest and highest shares of executives (2022)"
  )

# =========================
# 2016
# =========================

Merge |>
  filter(Annee == 2016, !is.na(code_departement)) |>
  group_by(code_departement, nom_departement) |>
  summarise(
    cadres = sum(`Pop 15 ans ou plus Cadres, Prof. intel. sup.`, na.rm = TRUE),
    denom  = sum(`Pop 15 ans ou plus`, na.rm = TRUE),
    TauxCSPsup = 100 * cadres / denom,
    .groups = "drop"
  ) |>
  filter(is.finite(TauxCSPsup), denom > 0) |>
  arrange(TauxCSPsup) -> d_2016

d_2016 |>
  slice_head(n = 10) |>
  transmute(
    `Lowest shares (2016)` = nom_departement,
    `CSP+ share` = round(TauxCSPsup, 2)
  ) -> low_2016

d_2016 |>
  slice_tail(n = 10) |>
  arrange(desc(TauxCSPsup)) |>
  transmute(
    `Highest shares (2016)` = nom_departement,
    `CSP+ share` = round(TauxCSPsup, 2)
  ) -> high_2016

bind_cols(low_2016, high_2016) |>
  kable(
    digits = 2,
    caption = "CSP+ — Departments with the lowest and highest shares of executives (2016)"
  )
```

#### Unemployment rate across departments

```{r, echo=FALSE, fig.width=14, fig.height=18}
chom_dep <- Merge |>
  filter(Annee %in% c(2016, 2022)) |>
  filter(!is.na(code_departement)) |>
  mutate(
    dep_label = paste0(code_departement, " – ", nom_departement),
    Annee = factor(Annee, levels = c("2016","2022")),
    pop2064 = `Pop Hommes 20-64 ans` + `Pop Femmes 20-64 ans`
  ) |>
  group_by(dep_label, Annee) |>
  summarise(
    chomeurs = sum(chomeur_total, na.rm = TRUE),
    pop2064  = sum(pop2064, na.rm = TRUE),
    taux_chom = 100 * chomeurs / pop2064,
    .groups = "drop"
  )


ggplot(chom_dep, aes(x = Annee, y = taux_chom, fill = Annee)) +
  geom_col(width = 0.8, position = position_dodge2(reverse = TRUE)) +
  coord_flip() +
  facet_wrap(~ dep_label, ncol = 3) +
  scale_fill_manual(
    values = c(
      "2016" = "#66C2A5",
      "2022" = "#FC8D62"
    )
  ) +
  labs(
    title = "Unemployment Rate by Department",
    subtitle = "Comparison between 2016 and 2022 (ratio of unemployed persons to the population aged 20–64).",
    x = NULL,
    y = "Unemployment rate (%)",
    fill = "Year"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold", size = 9),
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(size = 8),
    panel.grid.major.y = element_blank()
  )
```

```{r, echo=FALSE}
# =========================
# 2022
# =========================
Merge |>
  filter(Annee == 2022, !is.na(code_departement)) |>
  mutate(
    dep_label = paste0(code_departement, " – ", nom_departement),
    pop2064 = `Pop Hommes 20-64 ans` + `Pop Femmes 20-64 ans`
  ) |>
  group_by(dep_label) |>
  summarise(
    chomeurs = sum(chomeur_total, na.rm = TRUE),
    pop2064  = sum(pop2064, na.rm = TRUE),
    taux_chom = 100 * chomeurs / pop2064,
    .groups = "drop"
  ) |>
  arrange(taux_chom) -> d_2022

d_2022 |>
  slice_head(n = 10) |>
  transmute(
    `Lowest rates (2022)` = dep_label,
    `Rate (%)` = round(taux_chom, 2)
  ) -> low_2022

d_2022 |>
  slice_tail(n = 10) |>
  arrange(desc(taux_chom)) |>
  transmute(
    `Highest rates (2022)` = dep_label,
    `Rate (%)` = round(taux_chom, 2)
  ) -> high_2022

bind_cols(low_2022, high_2022) |>
  kable(
    caption = "Unemployment rate by department — 2022 (10 lowest vs 10 highest)",
    align = "l"
  )


# =========================
# 2016
# =========================
Merge |>
  filter(Annee == 2016, !is.na(code_departement)) |>
  mutate(
    dep_label = paste0(code_departement, " – ", nom_departement),
    pop2064 = `Pop Hommes 20-64 ans` + `Pop Femmes 20-64 ans`
  ) |>
  group_by(dep_label) |>
  summarise(
    chomeurs = sum(chomeur_total, na.rm = TRUE),
    pop2064  = sum(pop2064, na.rm = TRUE),
    taux_chom = 100 * chomeurs / pop2064,
    .groups = "drop"
  ) |>
  arrange(taux_chom) -> d_2016

d_2016 |>
  slice_head(n = 10) |>
  transmute(
    `Lowest rates (2016)` = dep_label,
    `Rate (%)` = round(taux_chom, 2)
  ) -> low_2016

d_2016 |>
  slice_tail(n = 10) |>
  arrange(desc(taux_chom)) |>
  transmute(
    `Highest rates (2016)` = dep_label,
    `Rate (%)` = round(taux_chom, 2)
  ) -> high_2016

bind_cols(low_2016, high_2016) |>
  kable(
    caption = "Unemployment rate by department — 2016 (10 lowest vs 10 highest)",
    align = "l"
  )
```

#### Relocation Rate across departments

```{r, echo=FALSE}
dem_dep <- Merge |>
  filter(Annee %in% c(2016, 2022)) |>
  filter(!is.na(code_departement)) |>
  mutate(
    dep_label = paste0(code_departement, " – ", nom_departement),
    Annee = factor(Annee, levels = c("2022","2016")),
    non_demenage = 
      `Pop 1 an ou plus habitant 1 an avt même logt` +
      `Pop 1 an ou plus habitant 1 an avt autre logt même commune`
  ) |>
  group_by(dep_label, Annee) |>
  summarise(
    non_demenage = sum(non_demenage, na.rm = TRUE),
    population   = sum(Population, na.rm = TRUE),
    taux_dem = 100 * (1 - non_demenage / population),
    .groups = "drop"
  )


ggplot(dem_dep, aes(x = Annee, y = taux_dem, fill = Annee)) +
  geom_col(width = 0.8) +
  coord_flip() +
  facet_wrap(~ dep_label, ncol = 3) +
  scale_fill_manual(
    values = c(
      "2016" = "#8DA0CB",  
      "2022" = "#E78AC3" 
    )
  ) +
  labs(
    title = "Residential Mobility Rate by Department",
    subtitle = "2016 vs 2022 Comparison (Resident Population)",
    x = NULL,
    y = "Relocation rate (%)",
    fill = "Year"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold", size = 9),
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(size = 8),
    panel.grid.major.y = element_blank()
  )
```

```{r, echo=FALSE}
# =========================
# 2022 — Mobility rate
# =========================
Merge |>
  filter(Annee == 2022, !is.na(code_departement)) |>
  mutate(
    non_demenage =
      `Pop 1 an ou plus habitant 1 an avt même logt` +
      `Pop 1 an ou plus habitant 1 an avt autre logt même commune`
  ) |>
  group_by(code_departement, nom_departement) |>
  summarise(
    non_demenage = sum(non_demenage, na.rm = TRUE),
    population   = sum(Population, na.rm = TRUE),
    taux_dem     = 100 * (1 - non_demenage / population),
    .groups = "drop"
  ) |>
  filter(is.finite(taux_dem), population > 0) |>
  arrange(taux_dem) -> d_2022

d_2022 |>
  slice_head(n = 10) |>
  transmute(
    `Lowest rates (2022)` = nom_departement,
    `Mobility rate (%)`   = round(taux_dem, 2)
  ) -> low_2022

d_2022 |>
  slice_tail(n = 10) |>
  arrange(desc(taux_dem)) |>
  transmute(
    `Highest rates (2022)` = nom_departement,
    `Mobility rate (%)`    = round(taux_dem, 2)
  ) -> high_2022

bind_cols(low_2022, high_2022) |>
  kable(
    digits = 2,
    caption = "Residential mobility — Departments with the lowest and highest rates (2022)",
    align = "l"
  )


# =========================
# 2016 — Mobility rate
# =========================
Merge |>
  filter(Annee == 2016, !is.na(code_departement)) |>
  mutate(
    non_demenage =
      `Pop 1 an ou plus habitant 1 an avt même logt` +
      `Pop 1 an ou plus habitant 1 an avt autre logt même commune`
  ) |>
  group_by(code_departement, nom_departement) |>
  summarise(
    non_demenage = sum(non_demenage, na.rm = TRUE),
    population   = sum(Population, na.rm = TRUE),
    taux_dem     = 100 * (1 - non_demenage / population),
    .groups = "drop"
  ) |>
  filter(is.finite(taux_dem), population > 0) |>
  arrange(taux_dem) -> d_2016

d_2016 |>
  slice_head(n = 10) |>
  transmute(
    `Lowest rates (2016)` = nom_departement,
    `Mobility rate (%)`   = round(taux_dem, 2)
  ) -> low_2016

d_2016 |>
  slice_tail(n = 10) |>
  arrange(desc(taux_dem)) |>
  transmute(
    `Highest rates (2016)` = nom_departement,
    `Mobility rate (%)`    = round(taux_dem, 2)
  ) -> high_2016

bind_cols(low_2016, high_2016) |>
  kable(
    digits = 2,
    caption = "Residential mobility — Departments with the lowest and highest rates (2016)",
    align = "l"
  )
```

```{r, echo=FALSE}
Merge |>
  filter(Annee %in% c(2016, 2022)) |>
  filter(!is.na(TauxChom), !is.na(Prixm2Moyen)) |>
  group_by(Annee) |>
  mutate(
    chom_cat = case_when(
      TauxChom <= quantile(TauxChom, 1/3, na.rm = TRUE) ~ "Low unemployment",
      TauxChom <= quantile(TauxChom, 2/3, na.rm = TRUE) ~ "Moderate unemployment",
      TRUE                                                ~ "High unemployment"
    )
  ) |>
  ungroup() -> df_box_chom
```

```{r, echo=FALSE}
df_box_chom |>
  ggplot(aes(x = factor(Annee), y = Prixm2Moyen, fill = factor(Annee))) +
  geom_boxplot() +
  facet_wrap(~ chom_cat, nrow = 1) +
  scale_fill_brewer(palette = "Set2")+
  theme_minimal()+
  labs(
    title = "Price distribution (€/m²) by unemployment group",
    x = NULL, y = "Price (€/m²)"
  )
```

```{r, echo=FALSE}
tab_dist_chom <- df_box_chom |>
  group_by(Annee, chom_cat) |>
  summarise(
    mean   = mean(Prixm2Moyen, na.rm = TRUE),
    se     = sd(Prixm2Moyen, na.rm = TRUE) / sqrt(sum(!is.na(Prixm2Moyen))),
    q1     = quantile(Prixm2Moyen, 0.25, na.rm = TRUE),
    median = median(Prixm2Moyen, na.rm = TRUE),
    q3     = quantile(Prixm2Moyen, 0.75, na.rm = TRUE),
    n      = sum(!is.na(Prixm2Moyen)),
    .groups = "drop"
  ) |>
  mutate(across(c(mean, se, q1, median, q3), ~ round(.x, 1))) |>
  arrange(chom_cat, Annee)

tab_dist_chom_wide <- tab_dist_chom |>
  select(Annee, chom_cat, mean, se, q1, median, q3) |>
  pivot_wider(
    names_from  = Annee,
    values_from = c(mean, se, q1, median, q3),
    names_glue  = "{.value}_{Annee}"
  ) |>
  arrange(chom_cat)

tab_dist_chom_wide |>
  kable(
    caption = "Price distribution summary (€/m²): 2016 vs 2022 within each unemployment group",
    col.names = c(
      "Unemployment group",
      "Mean 2016","Mean 2022",
      "SE 2016","SE 2022",
      "Q1 2016","Q1 2022",
      "Median 2016","Median 2022",
      "Q3 2016","Q3 2022"
    )
  )
```

```{r, echo=FALSE}
Merge |>
  filter(Annee %in% c(2016, 2022)) |>
  filter(!is.na(TauxCSPsup), !is.na(Prixm2Moyen)) |>
  group_by(Annee) |>
  mutate(
    csp_cat = case_when(
      TauxCSPsup <= quantile(TauxCSPsup, 1/3, na.rm = TRUE) ~ "Low CSP+",
      TauxCSPsup <= quantile(TauxCSPsup, 2/3, na.rm = TRUE) ~ "Moderate CSP+",
      TRUE                                                  ~ "High CSP+"
    )
  ) |>
  ungroup() -> df_box_csp
```

```{r, echo=FALSE}
df_box_csp |>
  ggplot(aes(x = factor(Annee), y = Prixm2Moyen, fill = factor(Annee))) +
  geom_boxplot() +
  facet_wrap(~ csp_cat, nrow = 1) +
  scale_fill_brewer(palette = "Set2")+
  theme_minimal()+
  labs(
    title = "Price distribution (€/m²) by CSP+ group",
    x = NULL, y = "Price (€/m²)"
  )
```

```{r, echo=FALSE}
tab_dist_csp <- df_box_csp |>
  group_by(Annee, csp_cat) |>
  summarise(
    mean   = mean(Prixm2Moyen, na.rm = TRUE),
    se     = sd(Prixm2Moyen, na.rm = TRUE) / sqrt(sum(!is.na(Prixm2Moyen))),
    q1     = quantile(Prixm2Moyen, 0.25, na.rm = TRUE),
    median = median(Prixm2Moyen, na.rm = TRUE),
    q3     = quantile(Prixm2Moyen, 0.75, na.rm = TRUE),
    n      = sum(!is.na(Prixm2Moyen)),
    .groups = "drop"
  ) |>
  mutate(across(c(mean, se, q1, median, q3), ~ round(.x, 1))) |>
  arrange(csp_cat, Annee)

tab_dist_csp_wide <- tab_dist_csp |>
  select(Annee, csp_cat, mean, se, q1, median, q3) |>
  pivot_wider(
    names_from  = Annee,
    values_from = c(mean, se, q1, median, q3),
    names_glue  = "{.value}_{Annee}"
  ) |>
  arrange(csp_cat)

tab_dist_csp_wide |>
  kable(
    caption = "Price distribution summary (€/m²): 2016 vs 2022 within each csp+ group",
    col.names = c(
      "csp+ group",
      "Mean 2016","Mean 2022",
      "SE 2016","SE 2022",
      "Q1 2016","Q1 2022",
      "Median 2016","Median 2022",
      "Q3 2016","Q3 2022"
    )
  )
```

```{r, echo=FALSE}
Merge |>
  filter(Annee %in% c(2016, 2022)) |>
  filter(!is.na(TauxDem), !is.na(Prixm2Moyen)) |>
  group_by(Annee) |>
  mutate(
    mob_cat = case_when(
      TauxDem <= quantile(TauxDem, 1/3, na.rm = TRUE) ~ "Low mobility",
      TauxDem <= quantile(TauxDem, 2/3, na.rm = TRUE) ~ "Moderate mobility",
      TRUE                                             ~ "High mobility"
    )
  ) |>
  ungroup() -> df_box_mob
```

```{r, echo=FALSE}
df_box_mob |>
  ggplot(aes(x = factor(Annee), y = Prixm2Moyen, fill = factor(Annee))) +
  geom_boxplot() +
  facet_wrap(~ mob_cat, nrow = 1) +
  scale_fill_brewer(palette = "Set2")+
  theme_minimal()+
  labs(
    title = "Price distribution by residential mobility group (2016 vs 2022)",
    x = NULL, y = "Price (€/m²)"
  )
```

```{r, echo=FALSE}
tab_dist_mob <- df_box_mob |>
  group_by(Annee, mob_cat) |>
  summarise(
    mean   = mean(Prixm2Moyen, na.rm = TRUE),
    se     = sd(Prixm2Moyen, na.rm = TRUE) / sqrt(sum(!is.na(Prixm2Moyen))),
    q1     = quantile(Prixm2Moyen, 0.25, na.rm = TRUE),
    median = median(Prixm2Moyen, na.rm = TRUE),
    q3     = quantile(Prixm2Moyen, 0.75, na.rm = TRUE),
    n      = sum(!is.na(Prixm2Moyen)),
    .groups = "drop"
  ) |>
  mutate(across(c(mean, se, q1, median, q3), ~ round(.x, 1))) |>
  arrange(mob_cat, Annee)

tab_dist_mob_wide <- tab_dist_mob |>
  select(Annee, mob_cat, mean, se, q1, median, q3) |>
  pivot_wider(
    names_from  = Annee,
    values_from = c(mean, se, q1, median, q3),
    names_glue  = "{.value}_{Annee}"
  ) |>
  arrange(mob_cat)

tab_dist_mob_wide |>
  kable(
    caption = "Price distribution summary (€/m²): 2016 vs 2022 within each mobility group",
    col.names = c(
      "Mobility group",
      "Mean 2016","Mean 2022",
      "SE 2016","SE 2022",
      "Q1 2016","Q1 2022",
      "Median 2016","Median 2022",
      "Q3 2016","Q3 2022"
    )
  )
```


# Econometric analysis on prices

## Implementing the baseline two-way fixed effect regression

We implement a two-way fixed effects (TWFE) panel regression at the municipal level for 2016 and 2022. The dependent variable is the logarithm of the average price per square meter. We express the dependent variable using the logarithm in order to interpret the coefficients (beta) as a percentage change in price. The main regressors reflect the local demographic and socioeconomic structure: proportion of executives (CSP+) (TauxCSPsup), unemployment rate (TauxChom), and residential mobility rate (TauxDem). 
We add real estate market controls: average surface area (SurfaceMoy), proportion of apartments (PropAppart, as a proxy of urban form/density) and market activity (ln_mut = log1p(Nb_mutations)). We include municipal fixed effects (absorbing local characteristics that are invariant over time, such as geography and amenities) and annual fixed effects (absorbing national shocks). Standard errors are pooled at the municipal level.

```{r}
Merge_reg <- Merge |>
  filter(Annee %in% c(2016, 2022)) |>
  mutate(
    ln_prix_m2 = log(Prixm2Moyen),
    ln_mut = log1p(Nb_mutations)
  ) |>
  filter(is.finite(ln_prix_m2))

m_twfe <- feols(
  ln_prix_m2 ~ TauxCSPsup + TauxChom + TauxDem + SurfaceMoy + PropAppart + ln_mut |
    INSEE_COM + Annee,
  data = Merge_reg,
  vcov = ~ INSEE_COM  # SE cluster commune
)

etable(m_twfe)
```

## First difference regression
To make the interpretation within change even more straightforward, we reshape the data into a wide format and calculate the differences between 2016 and 2022:
[\Delta \ln(P_{i}) = \ln(P_{i,2022}) - \ln(P_{i,2016}),] and we do the same for the regressors. We then perform a cross-sectional regression of price growth on changes in demographic composition and market conditions.

```{r}
Merge_wide <- Merge_reg |>
  select(INSEE_COM, Prixm2Moyen, TauxCSPsup, TauxChom, TauxDem,
         SurfaceMoy, PropAppart, Nb_mutations, Annee) |>
  tidyr::pivot_wider(
    names_from = Annee,
    values_from = c(Prixm2Moyen, TauxCSPsup, TauxChom, TauxDem,
                    SurfaceMoy, PropAppart, Nb_mutations, Annee),
    names_sep = "_"
  ) |>
  mutate(
    d_ln_prix = log(Prixm2Moyen_2022) - log(Prixm2Moyen_2016),
    d_csp = TauxCSPsup_2022 - TauxCSPsup_2016,
    d_chom = TauxChom_2022 - TauxChom_2016,
    d_dem = TauxDem_2022 - TauxDem_2016,
    d_surface = SurfaceMoy_2022 - SurfaceMoy_2016,
    d_propapp = PropAppart_2022 - PropAppart_2016,
    d_ln_mut = log1p(Nb_mutations_2022) - log1p(Nb_mutations_2016)
  ) |>
  filter(is.finite(d_ln_prix))

m_fd <- lm(d_ln_prix ~ d_csp + d_chom + d_dem + d_surface + d_propapp + d_ln_mut,
           data = Merge_wide)
summary(m_fd)

```

## Adding department-by-year fixed effects
Next, we strengthen the identification by replacing the simple fixed effects for the year with fixed effects for department × year. This allows us to absorb shocks common to all municipalities in the same department during a given year, such as local macroeconomic conditions or departmental policies. The remaining variation therefore lies within the municipality, excluding department-specific temporal shocks.

```{r}
m_depyear <- feols(
  ln_prix_m2 ~ TauxCSPsup + TauxChom + TauxDem + SurfaceMoy + PropAppart + ln_mut |
    INSEE_COM + code_departement^Annee,
  data = Merge_reg,
  vcov = ~ INSEE_COM
)

etable(m_twfe, m_depyear)

```

## Heterogeneity by urbanization terciles (interaction model)
To check whether demographic effects differ between urban and rural areas, we create an urbanisation indicator by dividing municipalities into tertiles of PropAppart. We then estimate a TWFE model in which TauxCSPsup, TauxChom and TauxDem interact with the urban tertile. This allows the marginal effect of each demographic variable to vary between municipalities with low, medium, and high proportions of flats, in other words between less dense and more dense areas. Due to collinearity, fixest removes some interaction terms.

```{r}
Merge_reg <- Merge_reg |>
  mutate(urb_tercile = ntile(PropAppart, 3))

m_het <- feols(
  ln_prix_m2 ~ (TauxCSPsup + TauxChom + TauxDem) * i(urb_tercile) +
    SurfaceMoy + ln_mut |
    INSEE_COM + Annee,
  data = Merge_reg,
  vcov = ~ INSEE_COM
)

etable(m_het)
```

## Explicit age structure (children vs seniors)
As our research question is about demographic composition, we add a simple block on age structure: the share of children and the share of elderly people (combining those aged 60-74, 75-89 and 90 and over). We estimate a TWFE model with the same controls and fixed effects as before.

```{r}
Merge_reg <- Merge_reg |>
  mutate(
    share_0_14 = (`Pop 0-14 ans`) / Population * 100,
    share_60p  = (`Pop 60-74 ans` + `Pop 75-89 ans` + `Pop 90 ans ou plus`) / Population * 100
  )

m_age <- feols(
  ln_prix_m2 ~ share_0_14 + share_60p + TauxCSPsup + TauxChom + TauxDem +
    SurfaceMoy + PropAppart + ln_mut |
    INSEE_COM + Annee,
  data = Merge_reg,
  vcov = ~ INSEE_COM
)

etable(m_age)
```


## Liquidity restriction (Nb_mutations ≥ 30)
Finally, we apply the TWFE benchmark model to a restricted sample of more liquid municipalities (Nb_mutations >= 30). The idea is to reduce the measurement error in Prixm2Moyen and ensure that the dependent variable reflects a reasonably active market.
```{r}
Merge_reg2 <- Merge_reg |> filter(Nb_mutations >= 30)

m_twfe_liq <- feols(
  ln_prix_m2 ~ TauxCSPsup + TauxChom + TauxDem + SurfaceMoy + PropAppart + ln_mut |
    INSEE_COM + Annee,
  data = Merge_reg2,
  vcov = ~ INSEE_COM
)

etable(m_twfe, m_twfe_liq)
```

